---
title: "Macho-Nice Guy Dichotomy"
author: "Gilad Sarusi"
format: 
  html:
    toc: true
    number-sections: true
    self-contained: true
execute:
  warning: false
  message: false
editor: visual
---

# HistWords repo

For our purposes, I use the same approach chosen by Tessa E. S. Charlesworth and colleagues—namely, the HistWords repository, which is available on [GitHub](https://github.com/williamleif/histwords?tab=readme-ov-file).

### What's going on?

HistWords is an open-source dataset and toolkit that provides word embeddings for different historical periods, trained on Google Books Ngram (and other corpora). Each decade (1800–1990) has its own 300-dimensional vector space representing the meaning of words in that era.

Think of it as a longitudinal semantic dataset—rather than tracking people over time, we track words across decades. Each vector encodes a word’s position in the shared “meaning space” of its time.

#### What can we do with it?

- **Inspect meaning:** find the nearest neighbors of a word in any decade.

- **Compare meanings over time:** compute cosine similarity for the same word across decades to track semantic drift.

- **Study cultural change:** analyze how a word’s associations (e.g., *macho*, *nice*) evolve by decade.

- **Compose phrases:** average vectors (e.g., *macho* + *guy*) to represent multi-word concepts.

- **Align embeddings:** apply sequential Procrustes alignment to make spaces from different decades comparable.

### What are we comparing?

In this project, I adopt an approach similar to Tessa et al. designed to avoid the **alignment problem** inherent in diachronic word embeddings—which makes direct comparison of raw cosine similarities between two embedding spaces (e.g., the 1910s and the 1920s) invalid.

Instead of comparing one decade directly with another, I compare words **within each decade**. In other words, for every decade I ask a simple question: is a given word closer to the *Macho* prototype or to the *Nice Guy* prototype?

To do this, the code builds a vector for each prototype in each decade and then gives every word a **single contrast score**. Positive scores mean the word is closer to the Macho side; negative scores mean it is closer to the Nice side. This lets me describe change across time without pretending that the raw geometry of one decade can be directly compared with the raw geometry of another.

More concretely, the current script does four things:

1.  it builds decade-specific *Macho* and *Nice Guy* vectors in two different ways: a **nudge** version and an **inject** version;
2.  it uses an anchored scoring method to estimate where each word falls on the Macho-versus-Nice contrast within that decade;
3.  it keeps only words that also appear in the Warriner ratings, so later comparisons can be tied to valence; and
4.  in the Attractiveness section, it compares two scoring workflows side by side: the original pairwise cosine approach and the newer anchored approach.

Nevertheless, following Tessa et al.’s guidance, applying **orthogonal Procrustes** rotation can still be useful as a robustness check or for future extensions involving cross-decade word-trajectory analyses.

::: callout-caution
## [`The following code is written using the ds4psych: Data Science for Psychology: Natural Language workflow (https://ds4psych.com). Therefore, the toolset used here differs from Tessa’s.`]{.smallcaps}
:::

```{r loadPackages, echo = FALSE, message=FALSE}
# ----------------------------------------------------
#Packages
library(tidyverse)
library(vroom)       # fast reading of large txts
library(stringr) # string manipulation
library(purrr) # functional programming
library(sweater) # for loading embeddings
library(quanteda.textmodels) # PLS
library(quanteda) # for text analysis
library(embedplyr) # for working with embeddingsembedding manipulation
library(knitr) # kable
library(pls) # PLS


```

```{r Datapaths, include=FALSE}

# --- 0) Paths  -------------------

decades <- seq(1800, 1990, by = 10)      
model_names <- setNames(paste0("eng.all_sgns.",decades), decades) # all English books (Google N-grams)


```

Notably, in the 2000s there was a change in N-gram sampling. Therefore, embeddings using the corpus from the 2000s onward are excluded.

```{r}
## Words norm - for valence computation: 


warriner_path <- "wordstim/allwrdnorms_warriner.csv"         
# columns: word, valence, arousal, dominance - valence is the average rating on a 9-point scale from 1 (very negative) to 9 (very positive), obtained from Warriner and colleagues (2013)

# Theoretically we need to cosider that the fact that this was a crowdslabeling procedure taken place in 2013, the chatechristics of each trait in term of valence, arousal and dominance are reflecting the meaning of the word in 2013, which might be different from the meaning of the word in 1800. However, since we are interested in the relative valence of traits (e.g., is "kind" more positive than "dominant"?), and since these relative valence ratings are likely to be more stable over time than absolute ratings, I will proceed with using Warriner's ratings as a proxy for trait valence across all decades. This is a common approach in historical semantics research, although it is important to acknowledge this limitation when interpreting results.

```

As an alternative, we might consider NRC VAD (S. M. Mohammad, 2018), which contains 20,007 words with ratings between 0 and 1 for valence, arousal, and dominance.

```{r}

## List of traits:

traits_path <- "wordstim/traitlist.txt"

# We would latter on create a file storing those words for the code to be more elegant :

nice_guy_words <- c(
  "kind", "gentle", "caring", "sensitive", "considerate", "respectful",
  "supportive", "understanding", "warm", "affectionate",
  "polite", "reliable", "faithful", "honest", "thoughtful",
  "attentive", "humble", "nice","whipped", "romantic"
)
macho_guy_words <- c(
   "dominant", "assertive", "confident", "tough", "strong", "decisive",
  "bold", "powerful", "competitive", "aggressive", "independent", "unemotional",
  "fearless","alpha", "commanding", "ambitious", "macho"
)

```

### Two ways of defining the constructs

To make sure these trait lists are interpreted in a masculine context, I use two slightly different strategies.

The first is to **inject male anchors**. This means I add words such as *man*, *men*, *male*, and *guy* to each list before averaging the embeddings. The goal is to pull the resulting vector toward a masculine semantic context rather than a generic trait context. This is implemented in the *Concepts* chunk.

- **Cons / watch-outs:**\
  *Anchor dominance risk*—male anchors can outweigh trait terms if the lists are small. Moreover, shared anchors in both groups reduce the distance between group centroids (smaller margins).\
  **Monitor:** anchor share in each centroid (≤30–40%), cosine margin between group means, and Top-N stability across decades (overlap %).

The second approach, implemented in the next chunk, is a **masculinity projection nudge**. Here I build a male-versus-female direction in the embedding space and then shift each construct slightly toward the male side before scoring it. This is a softer adjustment than injection: instead of adding anchor words into the list itself, it gently pushes the final vector toward a masculine interpretation.

- **Cons / watch-outs:**\
  *Axis quality*—if the gender sets are unbalanced or drift strongly by decade, the direction can be noisy.\
  *Leakage*—over-weighting (large *w*) can wash out original semantics.\
  **Monitor:** axis stability across decades (cosine between decade-specific *d*<sub>male</sub>), use small *w* (e.g., 0.1–0.3) and run sensitivity analyses; compare results with and without the tweak.

Using both approaches gives me a built-in robustness check. If the results tell the same story under both methods, that makes the interpretation more convincing.

### Masculinity projection nudge {#musculinity-nudge}

To interpret these trait bundles as referring to men rather than to personality in general, I create a female-to-male direction in the semantic space and add part of that direction to the construct vector. In plain language, this step nudges the vector so that later comparisons are more likely to reflect the masculine reading of the traits.

```{r}



# E: numeric matrix (rows = tokens, cols = dims), 

# male_tokens / female_tokens: anchors for the axis (can leave defaults)

 male_anchors <-  c("man","men","male","guy")  #Question: should we add 'boy'? 
 female_anchors <-  c("woman", "women", "female","girl")

 
 build_male_female_axis <- function(embeddings) {
  male_ddr <- predict(embeddings, male_anchors) |> average_embedding()
  female_ddr <- predict(embeddings, female_anchors) |> average_embedding()
  male_ddr - female_ddr
}

 

# DDR-based masculinity function:

# w = nudge strength (0 = no nudge, >0 = push toward masculine direction)
 
# default is set to 1 to inteperate results easily and to make sure that the wors be are assigning is regards to word_GUY not that word in general
 
 
masculinity_nudge <- function(ddr_vec, # Or any dim1:dim300 embedding object
                              E, # The embedding space 
                              w = 1) {
 axis <- build_male_female_axis(E)
  axis <- as.numeric(axis)
  
  if (is.matrix(ddr_vec)) {
    
    ddr_vec_nudged <- sweep(
      ddr_vec,
      2,
      w * axis,
      FUN = "+"
    )
    
  } else {
    
    ddr_vec <- as.numeric(ddr_vec)
    ddr_vec_nudged <- ddr_vec + w * axis
    
  }
  
  ddr_vec_nudged

}



```

The second approach, the **Inject approach**, does the same job more directly: it adds male anchor words to each word list before averaging, so the resulting construct vector is explicitly tied to a masculine context.

```{r Concepts, include=FALSE}
# --- 2) Defining concepts (labels) -----------------------------


# Inject them inside the Macho/Nice Construct 
nice_guy_words_injected <- nice_guy_words |>
  c(male_anchors)

macho_guy_words_injected <- macho_guy_words |>
    c(male_anchors)

```

```{r loadData1, include=FALSE}
# --- 3) Loading Data  ----------------


## -------------  Warriner's valence-----------

#(I'm keeping Arousal and Dominance since it might be useful later)

warriner <- read_csv(warriner_path, show_col_types = FALSE) |>
  transmute(word=tolower(Word),
         #Valence
         valence=V.Mean.Sum,
         # Arousal
         arousal=A.Mean.Sum,
         # Dominance
         dominance = D.Mean.Sum,
         id = as.character(row_number())
         ) |>
  
  # dealing with duplicate: averageing across duplicate
  group_by(word) |>
  summarise(
    valence = mean(valence, na.rm = TRUE),
    arousal = mean(arousal, na.rm = TRUE),
    dominance = mean(dominance, na.rm = TRUE),
    id = first(id)
  ) 


## ------------ Lists of traits --------------


trait_list<- read.delim(traits_path, header = FALSE) |>
  # the traits are stored in a column named V1 and I want to use it as a vector 
  pull(V1) 


  # I would fetch only the traits that got crowdslebelling in warriner: 
traits_warriner <- warriner |>
  filter(word %in% trait_list) 
  
```

### ENGall Model

Properties of ENGall_model:

- n of words: 100,000 (same word list across all decades). the authors only computed embeddings for words that were among the top 100,000 most frequent words over all time (for EngAll)
  - n of dimensions: 300.
  - Type of Embedding : SGNS.
  - Context Window : symmetric context window of 4 words.

```{r}
## ----------- ENGall model -----------

if (file.exists("engall/ENGall_model.RDS")) {
  ENGall_model<-readRDS("engall/ENGall_model.RDS")
}else {
  ENGall_model <- lapply(model_names, load_embeddings, save = FALSE) # list of embedding models
  saveRDS(ENGall_model, 
          file="engall/ENGall_model.RDS")
}
length(ENGall_model)
```

The ENGall model provides embeddings for the 100,000 most frequent words **across** all time. This means words coined later in history will be missing in earlier slices, and some words will be absent in particular decade models. The model is stored as a list of 20 elements, one per decade. In some decade data frames, certain words have no embedding.

let's map those words.

```{r}

#### ---- Check unavailable words by decade ----
n_avwords <- vector()
n_unav<- vector()
prop_avail <- vector()
n_total <- vector()

# Creating a list of available word and unavailable words: 

for (i in seq_along(ENGall_model)) {
  # Count available words as those with V1 != 0
  n_avwords[i] <- sum(ENGall_model[[i]][, 1] != 0)
  n_total[i] <- ENGall_model[[i]]|>
      nrow()  
  n_unav[i]  <- n_total[i] - n_avwords[i]
  prop_avail[i] <- n_avwords[i] / n_total[i]
}


avwords_n_decade_df <- data.frame(
  decade      = decades,
  n_avwords   = n_avwords,
  n_unavwords = n_unav,
  prop_avail  = prop_avail
)

```

```{r availWordsTable}
kable(avwords_n_decade_df |>
        #adding comma
        mutate(across(where(is.numeric),scales::comma)),
      digits = 3,
      caption = "Proportion of Words with Embedding")

```

Out of 100,000 words, each decade has on average 36,347 available words.

Notably, the 1990s slice has \~71% of words available. This does **not** mean “29% are post-1990 words.” Rather, 29% of the top-100k (over 1800–1999) simply lack enough occurrences in the 1990s to train vectors—often because they are earlier-era words that faded out, or because they are rare or orthographically different in that decade.

Later, we will pull valence from Warriner’s list (\~14,000 words), which bounds our effective vocabulary to those items. Because not all Warriner words appear in ENGall, the intersection may be smaller.

Another concern is whether some of the words used to define our constructs are missing from the model.

Let's check!

```{r}

### ---- Check if macho/nice words by decade ------- 
# 1) Availability (token x decade)

avail <- map2_df(ENGall_model, decades, ~
  tibble(
    decade    = .y,
    token     = tolower(rownames(.x)),
    available = .x[, 1] != 0
  )
)

#  Define the sets you care about
sets <- list(
  macho = macho_guy_words,
  nice  = nice_guy_words,
  male = male_anchors
)

wanted <- enframe(sets, name = "set", value = "token") |>
  unnest(token) |>
  mutate(token = tolower(token))

# proportion per decade + list of missing words 
macho_nice_words_by_decade_df <- wanted |>
  left_join(avail, by = "token") |>
  mutate(available = tidyr::replace_na(available, FALSE)) |>
  group_by(set, decade) |>
  summarise(
    n_of_words_examined = n(),
    n_available = sum(available),
    prop_avail    = n_available / n_of_words_examined,
    missing     = list(token[!available]),
    .groups = "drop"
  ) |>
  arrange(set, decade)

```

```{r}

kable(macho_nice_words_by_decade_df,
      caption= "Construct's word presence across time")
```

It seems the words *cocky* and *macho* do not appear in ENGall at all. I therefore omit them so that the construct vectors are built only from words that actually exist in the model.

This leaves the Macho and Nice lists with slightly different lengths, which is worth keeping in mind when interpreting the results.

```{r}
length(macho_guy_words)
#[1] 20

length(macho_guy_words_injected)

#[1] 24

length(male_anchors)
# [1] 4

macho_guy_words <- setdiff(macho_guy_words,c("cocky", "macho"))
macho_guy_words_injected <- setdiff(macho_guy_words_injected,c("cocky", "macho"))

# Sanity check
length(macho_guy_words)
# [1] 18

length(macho_guy_words_injected)

# [1] 20

# OK I'm sane
    
    
```

# DDR


I next build the decade-specific semantic representations for *Macho* and *Nice Guy*. In this document I use the term **DDR** simply to mean the vector that summarizes one construct in one decade.

The code creates four such vectors for each decade:

- Macho, using the nudge approach
- Nice Guy, using the nudge approach
- Macho, using the inject approach
- Nice Guy, using the inject approach

## Pairwise

The first scoring method is the *Pairwise Cosine-Delta* method. For every word in the decade-specific embedding space, cosine similarity is computed separately to the Macho DDR and to the Nice DDR. The final score is then defined as:

cosine_delta = cosine(word, Macho DDR) - cosine(word, Nice DDR)

Thus, positive values indicate that a word is more similar to the *Macho than to the Nice*, whereas negative values indicate that a word is more similar to the *Nice than to the Macho*. This method treats the two constructs as two separate reference points and asks whether a word is relatively closer to one than the other- non Dichatomic.

## Anchored Method

The second scoring method is the anchored Macho–Nice axis. Here, the Macho DDR is passed as the positive anchor and the Nice DDR is passed as the negative anchor using

`get_sims(..., method = "anchored").` Rather than computing two independent similarities and subtracting them, the anchored method places each word on a bipolar semantic continuum defined by the Nice pole at one end and the Macho pole at the other. Conceptually, this method asks where each word falls on the semantic axis:

Nice ←———— word ————→ Macho

This scale is ranging form 0 (nice) to 1 (macho). Higher anchored scores indicate a position closer to the Macho pole, whereas lower scores indicate a position closer to the Nice pole. This method therefore treats Macho and Nice as two ends of a single semantic dimension and therefore dichatomic.

## PLS
The third scoring method is the PLS-based Macho–Nice axis. In this approach, a supervised Partial Least Squares model is trained to distinguish Macho seed words from Nice seed words. Macho words are coded as +1, and Nice words are coded as -1. The resulting one-component PLS projection is used as a supervised semantic axis that maximally separates the two word groups. Each word in the decade-specific embedding space is then scored by its cosine similarity to this PLS-derived axis. Positive PLS scores indicate alignment with the Macho-oriented side of the supervised axis, whereas negative scores indicate alignment with the Nice-oriented side.

The three methods therefore answer related but non-identical questions. The pairwise cosine-delta method asks whether a word is more similar to Macho than to Nice. The anchored method asks where a word lies on the bipolar Nice–Macho continuum. The PLS method asks how strongly a word aligns with a supervised axis learned from the Macho versus Nice seed-word distinction. Because these methods produce scores using different geometric procedures, their raw magnitudes should not be interpreted as directly equivalent.

Instead, they should be compared in terms of directional agreement, rank-order similarity, overlap in high-scoring and low-scoring words, and associations with external lexical variables.
```{r pls-helper}
make_pls_macho_nice_axis <- function(embeddings,
                                     macho_words,
                                     nice_words,
                                     masculine_nudge = FALSE) {
  
  train_words <- tibble(
    word = c(macho_words, nice_words),
    y = c(
      rep(1, length(macho_words)),
      rep(-1, length(nice_words))
    )
  ) |>
    distinct(word, .keep_all = TRUE) |>
    filter(word %in% rownames(embeddings))

  X <- predict(embeddings, train_words$word)

  if (masculine_nudge) {
    X <- masculinity_nudge(X, E = embeddings)
  }

  X <- as.matrix(X)

  keep <- complete.cases(X) & rowSums(X^2, na.rm = TRUE) > 0

  X <- X[keep, , drop = FALSE]
  y <- train_words$y[keep]

  if (length(y) != nrow(X)) {
    stop("PLS training mismatch: X rows and y labels do not match.")
  }

  if (length(unique(y)) < 2) {
    stop("Only one class remains after removing missing embeddings.")
  }

  pls_data <- as.data.frame(X)
  pls_data$y <- y

  pls_fit <- plsr(
    y ~ .,
    data = pls_data,
    ncomp = 1,
    scale = FALSE,
    validation = "none"
  )

  axis <- as.numeric(pls_fit$projection[, 1])
  scores <- as.numeric(X %*% axis)

  if (mean(scores[y == 1]) < mean(scores[y == -1])) {
    axis <- -axis
  }

  axis / sqrt(sum(axis^2))
}
```

```{r DDRs,warning=FALSE}

# PlS
pls_nudge_axis_by_decade <- vector("list", length(ENGall_model))
pls_inject_axis_by_decade <- vector("list", length(ENGall_model))

Macho_DDR_nudge_by_decade <- vector("list", length(ENGall_model))
Nice_DDR_nudge_by_decade <- vector("list", length(ENGall_model))

Macho_DDR_injected_by_decade <- vector("list", length(ENGall_model))
Nice_DDR_injected_by_decade <- vector("list", length(ENGall_model))

decade_scores <- vector("list", length(ENGall_model))

for (i in seq_along(ENGall_model)) {
  embeddings <- ENGall_model[[i]]
  embeddings <- embeddings[magnitude(embeddings) != 0, ]
  
  # Nudged approach
  Nice_DDR_nudge_by_decade[[i]] <- masculinity_nudge(
    ddr_vec = average_embedding(predict(embeddings, nice_guy_words)),
    E = embeddings
  ) 
  
  Macho_DDR_nudge_by_decade[[i]] <- masculinity_nudge(
    ddr_vec = average_embedding(predict(embeddings, macho_guy_words)),
    E = embeddings
  ) 
  
  # Inject approach
  Nice_DDR_injected_by_decade[[i]] <- average_embedding(
    predict(embeddings, nice_guy_words_injected)
  )
  Macho_DDR_injected_by_decade[[i]] <- average_embedding(
    predict(embeddings, macho_guy_words_injected) 
  )
  
  # Method 1: Pairwise Deltas 
  cosine_scores_word_level <- embeddings |>
    get_sims(
      list(
        "macho_cos_nudge" = Macho_DDR_nudge_by_decade[[i]],
        "nice_cos_nudge" = Nice_DDR_nudge_by_decade[[i]],
        "macho_cos_inject" = Macho_DDR_injected_by_decade[[i]],
        "nice_cos_inject" = Nice_DDR_injected_by_decade[[i]]
      ), 
      method = "cosine"
    ) |> 
    mutate(
      cosine_delta_nudge = macho_cos_nudge - nice_cos_nudge,
      cosine_delta_inject = macho_cos_inject - nice_cos_inject
    ) |> 
    select(
      doc_id,
      cosine_delta_nudge,
      cosine_delta_inject
    ) 
  
  # Method 2: Anchored Macho-Nice Axis
  anchored_scores_word_level <- embeddings |> 
    get_sims(
      list(
        "macho_vs_nice_nudge" = list(
          pos = Macho_DDR_nudge_by_decade[[i]],
          neg = Nice_DDR_nudge_by_decade[[i]]
        ), 
        "macho_vs_nice_inject" = list(
          pos = Macho_DDR_injected_by_decade[[i]],
          neg = Nice_DDR_injected_by_decade[[i]]
        ) 
      ), 
      method = "anchored"
    )
  # Method 3: PLS
    pls_nudge_axis_by_decade[[i]] <- make_pls_macho_nice_axis(
    embeddings = embeddings,
    macho_words = macho_guy_words,
    nice_words = nice_guy_words,
    masculine_nudge = TRUE
  )
    
    pls_inject_axis_by_decade[[i]] <- make_pls_macho_nice_axis(
      embeddings = embeddings,
      macho_words = macho_guy_words_injected,
      nice_words = nice_guy_words_injected,
      masculine_nudge = FALSE
    )
    
    pls_scores_word_level <- embeddings |>
      get_sims(
        list(
          "pls_nudge"  = pls_nudge_axis_by_decade[[i]],
          "pls_inject" = pls_inject_axis_by_decade[[i]]
        ),
        method = "cosine"
      ) 
    
  decade_scores[[i]] <- anchored_scores_word_level |>
    left_join(cosine_scores_word_level,
              by = "doc_id")|>
    left_join(
      pls_scores_word_level |>
        select(
          doc_id,
          pls_nudge,
          pls_inject
        ),
      by = "doc_id"
    )
}

# Make it readable
names(Macho_DDR_nudge_by_decade) <- decades
names(Nice_DDR_nudge_by_decade) <- decades
names(Macho_DDR_injected_by_decade) <- decades
names(Nice_DDR_injected_by_decade) <- decades
names(decade_scores) <- decades


```

## Warriner List {#warriner}

Next, I keep **only** the words that also appear in the Warriner norms, because those are the words for which I have valence ratings.

In practice, this means I take the full set of anchored scores from the embedding model and trim it down to the subset of words that can later be interpreted in emotional terms.

```{r WarrinerOnly)}
# Starting with empty 20 long list 

warriner_with_scores <- vector("list", length(decade_scores))

for (i in seq_along(decade_scores)) {

  ds <- decade_scores[[i]]

  # --- standardize the token column name to "word" ---
  
  if ("doc_id" %in% names(ds)) {
    ds2 <- ds |> rename(word= doc_id)
  } else {
    stop(
      "Could not find a word column in decade_scores[[i]]. Check names(decade_scores[[i]])."
      )
  }

  # --- join: keep only Warriner words that exist in this decade vocab ---
  warriner_with_scores[[i]] <-
    warriner |>
    dplyr::inner_join(
      ds2,
      by = "word"
    )
}
names(warriner_with_scores) <- decades

warriner_with_scores


#-------- NOTICE! ------

# Before the comparisons I'll omit the words that were used to create the DDR, namely, macho_guy_word and nice_guy_words, to reduce noise .

omit_words <- unique(c(
  macho_guy_words, nice_guy_words,
  macho_guy_words_injected, nice_guy_words_injected
))

warriner_scores_omit <- lapply(warriner_with_scores, function(df) {
  df |> filter(!word %in% omit_words)
})


```



# RQs
## Q1

**Is there, overall, a difference in association across 200 years between Macho and Nice Guy?**

Here, positive values mean relatively stronger Macho association and negative values mean relatively stronger Nice association. So here we'll see whether the delta, anchored pole loading and pls changes, first overall and then by decade.




```{r compare-methods-prepare-long}
method_scores_long <- bind_rows(warriner_scores_omit, .id = "decade") |>
  mutate(
    decade = factor(decade, levels = as.character(decades)),
    decade_num = as.numeric(decade),

    # Anchored scores from get_sims(method = "anchored") are 0-1:
    # 0 = Nice pole, .5 = midpoint, 1 = Macho pole.
    # Center them so 0 = midpoint, positive = Macho, negative = Nice.
    anchored_nudge  = 2 * macho_vs_nice_nudge - 1,
    anchored_inject = 2 * macho_vs_nice_inject - 1,

    pairwise_nudge  = cosine_delta_nudge,
    pairwise_inject = cosine_delta_inject,

    pls_nudge  = pls_nudge,
    pls_inject = pls_inject
  ) |>
  select(
    word, id, valence, arousal, dominance,
    decade, decade_num,
    anchored_nudge, anchored_inject,
    pairwise_nudge, pairwise_inject,
    pls_nudge, pls_inject
  ) |>
  pivot_longer(
    cols = c(
      anchored_nudge, anchored_inject,
      pairwise_nudge, pairwise_inject,
      pls_nudge, pls_inject
    ),
    names_to = "score_type",
    values_to = "score"
  ) |>
  mutate(
    method = case_when(
      str_detect(score_type, "^anchored") ~ "Anchored",
      str_detect(score_type, "^pairwise") ~ "Pairwise delta",
      str_detect(score_type, "^pls") ~ "PLS"
    ),
    approach = case_when(
      str_detect(score_type, "nudge$") ~ "Nudge",
      str_detect(score_type, "inject$") ~ "Inject"
    ),
    direction = case_when(
      score > 0 ~ "Macho",
      score < 0 ~ "Nice",
      TRUE ~ "Neutral"
    ),
    valence_centered = valence - 5
  ) |>
  filter(is.finite(score))

```

## Attractiveness

This section intentionally keeps **both** Attractiveness workflows in the same document.

The older workflow, which came from `main`, asks two separate questions for each decade: "How close is Attractiveness to Macho?" and "How close is Attractiveness to Nice?" The newer workflow, from `anlysis-after-opt`, asks a single contrast question: "Is Attractiveness closer to Macho or to Nice?"

In principle, both approaches should tell a very similar story. Because they currently do not give identical results, I present them side by side and then compare them directly at the end.

### Word list

We begin by specifying a small list of adjectives that directly express attractiveness, such as *attractive*, *beautiful*, *handsome*, and *sexy*. As above, two versions are created:

- Nudge version: uses only the attractiveness descriptors.

- Inject version: includes the attractiveness descriptors plus the male anchor words, ensuring a stronger masculine reference frame in the embedding space.

This parallels the earlier Macho/Nice pipeline for consistency.

```{r attractiveness-words}
attractiveness_words <- c(
  "attractive",
  "beautiful",
  "handsome",
  "gorgeous",
  "pretty",
  "sexy",
  "cute",
  "alluring",
  "charming",
  "appealing"
)

attractiveness_words_injected <- attractiveness_words |>
  c(male_anchors)
```

For each decade, I build an Attractiveness vector using the same two strategies used above for Macho and Nice:

**Inject condition**

1.  Retrieve the embeddings of all attractiveness words + male anchors.
2.  Compute their **average embedding**, yielding the *Attractiveness Injected DDR* for that decade.

**Nudge condition**

1.  Retrieve the embeddings of the attractiveness adjectives.
2.  Average them to obtain the raw *Attractiveness Nudge DDR*.
3.  Apply the **`masculinity_nudge()`** transformation to align the DDR with the nudge paradigm used for the Macho/Nice DDRs.

```{r attractiveness-ddrs}
Attractiveness_DDR_injected_by_decade <- vector("list", length(ENGall_model))
Attractiveness_DDR_nudge_by_decade <- vector("list", length(ENGall_model))

for (i in seq_along(ENGall_model)) {
  embeddings <- ENGall_model[[i]]
  embeddings <- embeddings[magnitude(embeddings) != 0, ]
  
  Attractiveness_DDR_injected_by_decade[[i]] <- predict(
    embeddings,
    attractiveness_words_injected
  ) |>
    average_embedding()
  
  Attractiveness_DDR_nudge_by_decade[[i]] <- predict(
    embeddings,
    attractiveness_words
  ) |>
    average_embedding() |>
    masculinity_nudge(E = embeddings)
}

names(Attractiveness_DDR_injected_by_decade) <- decades
names(Attractiveness_DDR_nudge_by_decade) <- decades
```

### Original pairwise approach from main

This reproduces the original logic. For every decade and for each approach, it calculates:

1.  the cosine similarity between Attractiveness and Macho; and
2.  the cosine similarity between Attractiveness and Nice.

So the plot shows two lines for each approach. A higher line means that construct is more semantically similar to Attractiveness in that decade. Afterward, I also compute a simple summary difference, `Macho - Nice`, so the result can be compared more directly with the newer anchored method.

```{r attractiveness-pairwise-cosine}

cosine_from_attractiveness <- vector("list", length(ENGall_model))


for (i in seq_along(ENGall_model)) {
  
  # For each decade, we take the embeddings and add the DDRs for Macho, Nice to the embedding space. This allows us to compute cosine similarities between these DDRs and any other word in the space.
  embeddings <- ENGall_model[[i]]
  embeddings <- embeddings[magnitude(embeddings) != 0, ]
  
  embeddings_plus <- rbind(
    as.matrix(embeddings),
    macho_ddr_nudge  = as.numeric(Macho_DDR_nudge_by_decade[[i]]),
    nice_ddr_nudge   = as.numeric(Nice_DDR_nudge_by_decade[[i]]),
    macho_ddr_inject = as.numeric(Macho_DDR_injected_by_decade[[i]]),
    nice_ddr_inject  = as.numeric(Nice_DDR_injected_by_decade[[i]])
  ) |>
    embedplyr::as.embeddings()
  
  
  sims <- embeddings_plus |>
    get_sims(
      list(
        attract_nudge  = Attractiveness_DDR_nudge_by_decade[[i]],
        attract_inject = Attractiveness_DDR_injected_by_decade[[i]]
      ),
      method = "cosine"
    )
  
  cosine_from_attractiveness[[i]] <- tibble(
    decade = decades[i],
    approach = c("Nudge", "Inject"),
    sims_macho_attract = c(
      sims$attract_nudge[sims$doc_id == "macho_ddr_nudge"],
      sims$attract_inject[sims$doc_id == "macho_ddr_inject"]),
    sims_nice_attract = c(
      sims$attract_nudge[sims$doc_id == "nice_ddr_nudge"],
      sims$attract_inject[sims$doc_id == "nice_ddr_inject"]
    )
  )
  
  pairwise_attractiveness_df <- bind_rows(cosine_from_attractiveness) |>
  mutate(
    pairwise_delta = sims_macho_attract - sims_nice_attract
  )
}
cosine_attract_nudge_df <- pairwise_attractiveness_df |>
  filter(approach == "Nudge") |>
  select(-approach)

cosine_attract_inject_df <- pairwise_attractiveness_df |>
  filter(approach == "Inject") |>
  select(-approach)

```

```{r attractiveness-pairwise-plot, echo=FALSE, fig.width=10, fig.height=7, fig.align='center'}
pairwise_attractiveness_long <- pairwise_attractiveness_df |>
  pivot_longer(
    cols = c(sims_macho_attract, sims_nice_attract),
    names_to = "construct",
    values_to = "cosine_similarity"
  ) |>
  mutate(
    construct = recode(
      construct,
      sims_macho_attract = "Macho",
      sims_nice_attract = "Nice"
    )
  )

ggplot(pairwise_attractiveness_long, aes(x = decade, y = cosine_similarity, color = construct)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.6) +
  facet_wrap(~ approach, ncol = 1) +
  labs(
    title = "Original pairwise cosine approach",
    x = "Decade",
    y = "Cosine similarity to Attractiveness",
    color = "Construct"
  ) +
  theme_minimal()
```

```{r attractiveness-pairwise-test}
pairwise_attractiveness_tests <- bind_rows(
  {
    tt <- t.test(
      cosine_attract_nudge_df$sims_macho_attract,
      cosine_attract_nudge_df$sims_nice_attract,
      alternative = "two.sided"
    )
    tibble(
      approach = "Nudge",
      t_value = unname(tt$statistic),
      df = unname(tt$parameter),
      p_value = tt$p.value,
      mean_macho_cosim = mean(cosine_attract_nudge_df$sims_macho_attract),
      mean_nice_cosim = mean(cosine_attract_nudge_df$sims_nice_attract),
      mean_pairwise_delta = mean(
        cosine_attract_nudge_df$sims_macho_attract -
          cosine_attract_nudge_df$sims_nice_attract
      )
    )
  },
  {
    tt <- t.test(
      cosine_attract_inject_df$sims_macho_attract,
      cosine_attract_inject_df$sims_nice_attract,
      alternative = "two.sided"
    )
    tibble(
      approach = "Inject",
      t_value = unname(tt$statistic),
      df = unname(tt$parameter),
      p_value = tt$p.value,
      mean_macho_cosim = mean(cosine_attract_inject_df$sims_macho_attract),
      mean_nice_cosim = mean(cosine_attract_inject_df$sims_nice_attract),
      mean_pairwise_delta = mean(
        cosine_attract_inject_df$sims_macho_attract -
          cosine_attract_inject_df$sims_nice_attract
      )
    )
  }
)

kable(
  pairwise_attractiveness_tests,
  caption = "Original pairwise cosine tests from main",
  digits = 3
)
```

### Anchored contrast approach from anlysis-after-opt

This approach puts the Macho, Nice, and Attractiveness vectors into a small temporary semantic space for each decade and then uses `get_sims()` to ask a single question: is Attractiveness closer to Macho or to Nice?

The result is one contrast score per decade. Positive values mean Attractiveness is closer to Macho than to Nice; negative values mean it is closer to Nice than to Macho.

```{r attractiveness-anchored-cosine}
attractiveness_anchored_df <- lapply(seq_along(ENGall_model), function(i) {
  
  # -------------------------------------- 
  # THIS PART WAS DONE ORIGINALLY TO CREATE A SMALL EMBEDDING SPACE FOR THE THREE DDRs, BUT IT MAKES THE COMARISON WITH THE ORIGINAL PAIRWISE METHOD LESS DIRECT. SO I COMMENTED IT OUT.
  
  # ddr_space <- rbind(
  #   #nudge 
  #   macho_ddr_nudge = as.numeric(Macho_DDR_nudge_by_decade[[i]]),
  #   nice_ddr_nudge = as.numeric(Nice_DDR_nudge_by_decade[[i]]),
  #   attract_ddr_nudge = as.numeric(Attractiveness_DDR_nudge_by_decade[[i]]),
  #   #inject
  #   macho_ddr_inject = as.numeric(Macho_DDR_injected_by_decade[[i]]),
  #   nice_ddr_inject = as.numeric(Nice_DDR_injected_by_decade[[i]]),
  #   attract_ddr_inject = as.numeric(Attractiveness_DDR_injected_by_decade[[i]])
  # )
  # 
  # ddr_space <- embedplyr::as.embeddings(ddr_space)
  
  # DDR_SPACE WAS PASSED AS E TO GET_SIMS()
    # -------------------------------------- 

  
  embeddings <- ENGall_model[[i]]
  embeddings <- embeddings[magnitude(embeddings) != 0, ]
  
  
  embeddings_plus <- rbind(
    as.matrix(embeddings),
    attract_ddr_nudge  = as.numeric(Attractiveness_DDR_nudge_by_decade[[i]]),
    attract_ddr_inject = as.numeric(Attractiveness_DDR_injected_by_decade[[i]])
  ) |>
    embedplyr::as.embeddings()
  
  
  sims <- embeddings_plus |>
    get_sims(
      list(
        "macho_vs_nice_nudge" = list(
          pos = Macho_DDR_nudge_by_decade[[i]],
          neg = Nice_DDR_nudge_by_decade[[i]]
        ),
        "macho_vs_nice_inject" = list(
          pos = Macho_DDR_injected_by_decade[[i]],
          neg = Nice_DDR_injected_by_decade[[i]]
        )
      ),
      method = "anchored"
    )

  tibble(
    decade = decades[i],
    anchored_delta_nudge =
      sims$macho_vs_nice_nudge[sims$doc_id == "attract_ddr_nudge"],
    
    anchored_delta_inject =
      sims$macho_vs_nice_inject[sims$doc_id == "attract_ddr_inject"]
  )
}) |>
  bind_rows()

attractiveness_anchored_long <- attractiveness_anchored_df |>
  pivot_longer(
    cols = c(anchored_delta_nudge, anchored_delta_inject),
    names_to = "approach",
    values_to = "anchored_delta"
  ) |>
  mutate(
    
    # The anchored method produces a contrast score that ranges from 0 to 1, where 0.5 means Attractiveness is equidistant from Macho and Nice. To make it more directly comparable to the original pairwise cosine deltas (which can range from negtive to possitive values), I rescale the anchored deltas by multiplying by 2 and then shifting by -1. This way, a value of 0.5 in the anchored method corresponds to the same point of balance as in the pairwise method.
    anchored_centered = 2 * anchored_delta - 1,
    approach = recode(
      approach,
      anchored_delta_nudge = "Nudge",
      anchored_delta_inject = "Inject")
  )
```

```{r attractiveness-anchored-plot, echo=FALSE, fig.width=10, fig.height=6, fig.align='center'}
ggplot(attractiveness_anchored_long, aes(x = decade, y = anchored_centered, color = approach)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey45") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.6) +
  labs(
    title = "Anchored Macho-vs-Nice contrast for Attractiveness",
    x = "Decade",
    y = "Anchored contrast: Macho - Nice",
    color = "Approach"
  ) +
  theme_minimal()
```

```{r attractiveness-anchored-test}
# thsoe the change in attractiveness over time on the macho-nice axis shift significantly 


    
```

::: callout-note
## What now

Stuff to need to take into consideration and to thourghly continue the analysis

method: 1- Currently the masculinity nudge is set to w=1, meaning that the the nudge is very very storng toward the masculine side. try maybe something more sable. \* should inspect show what happening when varying w

2- the Attractiveness DDR also is constructed using the same "masculine context" embedding both inject and nudge method (which is also set to w=1!) I suspect that the results showed in @warriner that almost all the words loaded greatly more to the macho side of the axis (as shown in *macho_vs_nice* columns) is due to that the attractiveness is also pushed too strongly to the masculinity side.

- should inspect what happend when attractiveness not in the "masculine side" non-inject and non-nudged

future:

3 - Attractiveness as a sexual properties- changing the word bank 4- euclidean distance between the macho and nice guy construct as a support of the existance of a split. \* since only relative distance is possible for comparison across decades, a method is needed to examing the change (distance ) between the constructs. I'm thinking of taking the delta distance of every word from our macho/nice constructs, compute a delta and average across ALL the words and see if this average delta changes across time/

5- what other axaminations would indicate such "stereo-type- Split"?
:::

```{r pls-cav-attractiveness-by-decade-simple}
# Score one vector against one axis using cosine, so the PLS scale stays
# bounded and easier to compare across decades.
project_vec <- function(vec, axis) {
  cosine_sim(as.numeric(vec), as.numeric(axis))
}

pls_nudge_axis_by_decade <- vector("list", length(ENGall_model))
pls_cav_attractiveness_df <- vector("list", length(ENGall_model))

for (i in seq_along(ENGall_model)) {
  pls_nudge_axis_by_decade[[i]] <- make_pls_macho_nice_axis(
    embeddings = ENGall_model[[i]],
    macho_words = macho_guy_words,
    nice_words = nice_guy_words,
    masculine_nudge = TRUE
  )

  pls_cav_attractiveness_df[[i]] <- tibble(
    decade = decades[i],
    pls_delta_nudge = project_vec(
      Attractiveness_DDR_nudge_by_decade[[i]],
      pls_nudge_axis_by_decade[[i]]
    )
  )
}

names(pls_nudge_axis_by_decade) <- decades
pls_cav_attractiveness_df <- bind_rows(pls_cav_attractiveness_df)
```

### Method comparison

The figure below now puts all three nudge-based methods on the same plot for Attractiveness: the original pairwise cosine delta, the anchored contrast, and the PLS score.

```{r attractiveness-method-comparisonP2}
methods_comparison_df <- bind_rows(
  pairwise_attractiveness_df |>
    filter(approach == "Nudge") |>
    transmute(
      decade = decade,
      score = pairwise_delta,
      method = "Pairwise delta"
    ),
  attractiveness_anchored_long |> filter(approach == "Nudge")|> 
    transmute(
      decade = decade,
      score = anchored_centered,
      method = "Anchored"
    ),
  pls_cav_attractiveness_df |>
    transmute(
      decade = decade,
      score = pls_delta_nudge,
      method = "PLS"
    )
) 

methods_comparison_summary <- methods_comparison_df |>
  group_by(method) |>
  summarise(
    mean_score = mean(score, na.rm = TRUE),
    sd_score = sd(score, na.rm = TRUE),
    .groups = "drop"
  )

kable(
  methods_comparison_summary,
  caption = "Attractiveness comparison across the three nudge-based methods",
  digits = 4
)
```

```{r method-comp, echo=FALSE, fig.width=10, fig.height=6, fig.align='center'}
ggplot(
  methods_comparison_df,
  aes(x = decade, y = score, color = method)
) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey45") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.2) +
  labs(
    title = "Attractiveness across pairwise delta, anchored, and PLS",
    x = "Decade",
    y = "Macho - Nice score",
    color = "Method"
  ) +
  theme_minimal()
```

Because the nudge and inject variants point to the same substantive conclusion in the tables above, the remaining sections focus on the **nudge** specification only. I leave the inject results in place as a robustness check, but narrowing the downstream analyses to one masculinity adjustment keeps the cross-method comparisons readable without changing the story.

# Research Questions

The next three subsections bring back the research-question logic from Section 2.2 of `v1.html`, but now each question is answered with three parallel nudge-based methods:

- `Pairwise delta`: `cos(word, Macho) - cos(word, Nice)`
- `Anchored`: the `get_sims(..., method = "anchored")` contrast
- `PLS`: cosine to the learned PLS Macho-versus-Nice axis

```{r rq-helpers}
#Helper - Comparison:
compare_DDRs<- function(df) {
  
  df |> 
    mutate(
      # Nudge
      delta_nudge = macho_nudge - nice_nudge,
      # Inject
      delta_inject = macho_inject - nice_inject,
    )
}


# Adding the variables to the list 
warriner_COSINE_DDR_Diff <- list() 
for (i in seq_along(warriner_COSINE_DDR_omit)) {
  warriner_COSINE_DDR_Diff[[i]]<- compare_DDRs(warriner_COSINE_DDR_omit[[i]])
}
names(warriner_COSINE_DDR_Diff) <- decades 

# Make it a long format for it to be easier to work with :


bind_with_decade <- function(dflist) {
  
  tibble(
    decade = names(dflist),
    df     = dflist
  ) |>
    mutate(df = map2(df, decade, ~ .x |>
                       mutate(decade = .y))) |>
    pull(df) |>
    bind_rows()
}

raw_long <- bind_with_decade(warriner_COSINE_DDR_Diff) |>
  mutate(
    
    # Make `decade` an ordered factor (sorted numerically)
    
    decade     = factor(decade, levels = sort(unique(as.integer(decade))) |> as.character())
  ) 
```

```{r nudge-word-level-scores}

# This function computes the cosine similarity between each row of a matrix and one vector.
# So if mat contains many word embeddings, and vec is a semantic direction such as a Macho–Nice axis, the function returns one cosine score per word.


warriner_nudge_method_scores <- vector("list", length(ENGall_model))

for (i in seq_along(ENGall_model)) {
  embeddings <- ENGall_model[[i]]
  embeddings <- embeddings[magnitude(embeddings) != 0, ]

  available_words <- warriner |>

    filter(word %in% rownames(embeddings)) |>
    pull(word) 
    
    
  
  # get the embedding of all the available words from warrier's list
  word_embeddings <- predict(embeddings, available_words)

  
  
  # Method 1: Pairwise delta  
  pairwise_scores <- tibble(
    word = available_words,
    pairwise_delta =
      row_cosine_scores(word_embeddings, Macho_DDR_nudge_by_decade[[i]]) -
      row_cosine_scores(word_embeddings, Nice_DDR_nudge_by_decade[[i]])
  )

  # Method 2: Anchored 
  # decade_scores is the output of get_sims() for each decade when method is set to 'anchored' :
  anchored_scores <- decade_scores[[i]] |> 
    
    rename(word = doc_id) |>
    # decade_scores contains both nudge and inject, we take only nudge:
    select(word, anchored = macho_vs_nice_nudge) 

  # Method 3: PLS
  pls_scores <- tibble(
    word = available_words,
    pls = row_cosine_scores(word_embeddings, pls_nudge_axis_by_decade[[i]])
  )

  warriner_nudge_method_scores[[i]] <- warriner |>
    inner_join(pairwise_scores, by = "word") |>
    inner_join(anchored_scores, by = "word") |>
    inner_join(pls_scores, by = "word") |>
    
    # Omit the words that were used to create the DDRs, to reduce noise in the analysis:
    filter(!word %in% omit_words) |>
    mutate(
      decade = decades[i]    
    )

}

names(warriner_nudge_method_scores) <- decades

warriner_nudge_method_scores_wide <- bind_rows(warriner_nudge_method_scores)

warriner_nudge_method_scores_long <- warriner_nudge_method_scores_wide |>
  pivot_longer(
    cols = c(pairwise_delta, anchored, pls),
    names_to = "method",
    values_to = "score"
  ) |>
  mutate(
    method = recode(
      method,
      pairwise_delta = "Pairwise delta",
      anchored = "Anchored",
      pls = "PLS"
    ),
    decade = factor(decade, levels = decades),
    valence_centered = valence - 5
  )
```

```{r rq-helpers}
one_sample_summary <- function(x) {
  x <- x[is.finite(x)] 

  n <- length(x)
  mean_score <- mean(x)
  sd_score <- if (n > 1) sd(x) else NA_real_
  se <- if (n > 1) sd_score / sqrt(n) else NA_real_

  if (n > 1 && is.finite(sd_score) && sd_score > 0) {
    tt <- t.test(x, mu = 0)
    t_value <- unname(tt$statistic)
    df <- unname(tt$parameter)
    p_value <- tt$p.value
    ci_low <- unname(tt$conf.int[1])
    ci_high <- unname(tt$conf.int[2])
    d_z <- mean_score / sd_score
  } else {
    t_value <- NA_real_
    df <- NA_real_
    p_value <- NA_real_
    ci_low <- NA_real_
    ci_high <- NA_real_
    d_z <- NA_real_
  }

  tibble(
    n = n,
    mean_score = mean_score,
    sd_score = sd_score,
    se = se,
    t_value = t_value,
    df = df,
    p_value = p_value,
    d_z = d_z,
    ci_low = ci_low,
    ci_high = ci_high,
    direction = case_when(
      mean_score > 0 ~ "Macho",
      mean_score < 0 ~ "Nice",
      TRUE ~ "Neutral"
    ),
    sig = case_when(
      is.na(p_value) ~ "No",
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      TRUE ~ "No"
    )
  )
}

summarise_contrast_by_item <- function(data, only_traits = FALSE) {
  if (only_traits) {
    data <- data |>
      filter(word %in% traits_warriner$word)
  }

  data |>
    group_by(method, word) |>
    group_modify(~ one_sample_summary(.x$score)) |>
    ungroup()
}

make_top_table <- function(summary_df, top_n = 10, min_n = 5) {
  bind_rows(
    summary_df |>
      filter(sig != "No", n >= min_n) |>
      group_by(method) |>
      arrange(desc(d_z), .by_group = TRUE) |>
      slice_head(n = top_n) |>
      mutate(side = "Macho", rank = row_number()),
    summary_df |>
      filter(sig != "No", n >= min_n) |>
      group_by(method) |>
      arrange(d_z, .by_group = TRUE) |>
      slice_head(n = top_n) |>
      mutate(side = "Nice", rank = row_number())
  ) |>
    ungroup() |>
    select(method, side, rank, word, n, d_z, p_value)
}

method_overlap <- function(top_df) {
  if (n_distinct(top_df$method) < 2) {
    return(tibble(
      method_1 = character(),
      method_2 = character(),
      overlap_n = integer(),
      jaccard = double()
    ))
  }

  method_pairs <- combn(unique(top_df$method), 2, simplify = FALSE)

  map_dfr(method_pairs, function(pair) {
    left_words <- top_df |>
      filter(method == pair[1]) |>
      pull(word)

    right_words <- top_df |>
      filter(method == pair[2]) |>
      pull(word)

    overlap_words <- intersect(left_words, right_words)
    union_n <- length(union(left_words, right_words))

    tibble(
      method_1 = pair[1],
      method_2 = pair[2],
      overlap_n = length(overlap_words),
      jaccard = if (union_n > 0) length(overlap_words) / union_n else NA_real_
    )
  })
}

get_top_valence_series <- function(data, only_traits = FALSE, top_n = 10) {
  if (only_traits) {
    data <- data |>
      filter(word %in% traits_warriner$word)
  }

  bind_rows(
    data |>
      group_by(method, decade, decade_num) |>
      arrange(desc(score), .by_group = TRUE) |>
      slice_head(n = top_n) |>
      summarise(mean_valence = mean(valence_centered, na.rm = TRUE), .groups = "drop") |>
      mutate(side = "Macho"),
    data |>
      group_by(method, decade, decade_num) |>
      arrange(score, .by_group = TRUE) |>
      slice_head(n = top_n) |>
      summarise(mean_valence = mean(valence_centered, na.rm = TRUE), .groups = "drop") |>
      mutate(side = "Nice")
  )
}

fit_valence_trends <- function(df) {
  df |>
    group_by(level, side, method) |>
    group_modify(~ {
      fit <- lm(mean_valence ~ decade_num, data = .x)
      fit_summary <- summary(fit)

      tibble(
        intercept = unname(coef(fit)[["(Intercept)"]]),
        slope = unname(coef(fit)[["decade_num"]]),
        p_value = fit_summary$coefficients["decade_num", "Pr(>|t|)"],
        r_squared = fit_summary$r.squared
      )
    }) |>
    ungroup() |>
    mutate(
      sig = case_when(
        p_value < 0.001 ~ "***",
        p_value < 0.01 ~ "**",
        p_value < 0.05 ~ "*",
        TRUE ~ "No"
      )
    )
}
```

## Q1

**Is there, overall, a difference in association across 200 years between Macho and Nice Guy?**

Here, positive values mean relatively stronger Macho association and negative values mean relatively stronger Nice association. So for each method I test whether the mean score differs from zero, first overall and then by decade.

```{r rq1}
rq1_overall <- warriner_nudge_method_scores_long |>
  group_by(method) |>
  group_modify(~ one_sample_summary(.x$score)) |>
  ungroup()

rq1_by_decade <- warriner_nudge_method_scores_long |>
  group_by(method, decade) |>
  group_modify(~ one_sample_summary(.x$score)) |>
  ungroup()

rq1_by_decade_summary <- rq1_by_decade |>
  group_by(method) |>
  summarise(
    significant_decades = sum(sig != "No", na.rm = TRUE),
    mean_of_decade_means = mean(mean_score, na.rm = TRUE),
    min_decade_mean = min(mean_score, na.rm = TRUE),
    max_decade_mean = max(mean_score, na.rm = TRUE),
    .groups = "drop"
  )

rq1_method_agreement <- map_dfr(
  combn(c("pairwise_delta", "anchored", "pls"), 2, simplify = FALSE),
  function(cols) {
    x <- warriner_nudge_method_scores_wide[[cols[1]]]
    y <- warriner_nudge_method_scores_wide[[cols[2]]]

    tibble(
      method_1 = recode(
        cols[1],
        pairwise_delta = "Pairwise delta",
        anchored = "Anchored",
        pls = "PLS"
      ),
      method_2 = recode(
        cols[2],
        pairwise_delta = "Pairwise delta",
        anchored = "Anchored",
        pls = "PLS"
      ),
      correlation = cor(x, y, use = "complete.obs"),
      sign_agreement = mean(sign(x) == sign(y), na.rm = TRUE)
    )
  }
)

kable(
  rq1_overall,
  caption = "RQ1 overall test against zero for all three nudge-based methods",
  digits = 4
)

kable(
  rq1_by_decade_summary,
  caption = "RQ1 decade-level summary by method",
  digits = 4
)

kable(
  rq1_method_agreement,
  caption = "RQ1 agreement between scoring methods",
  digits = 4
)
```

```{r rq1-plot, echo=FALSE, fig.width=10, fig.height=6, fig.align='center'}
ggplot(rq1_by_decade, aes(x = as.numeric(as.character(decade)), y = mean_score, color = method)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey45") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  labs(
    title = "RQ1: decade-level Macho-minus-Nice scores",
    x = "Decade",
    y = "Mean score",
    color = "Method"
  ) +
  theme_minimal()
```

## Q2

**Across all decades, which words or traits are most associated with each construct?**

For the cross-decade ranking I require an item to appear in at least five decades. This keeps the overall tables from being driven by one- or two-decade accidents.

```{r rq2}
rq2_words_summary <- summarise_contrast_by_item(warriner_nudge_method_scores_long)
rq2_traits_summary <- summarise_contrast_by_item(
  warriner_nudge_method_scores_long,
  only_traits = TRUE
)

rq2_top_words <- make_top_table(rq2_words_summary, top_n = 10, min_n = 5)
rq2_top_traits <- make_top_table(rq2_traits_summary, top_n = 10, min_n = 5)

rq2_method_overlap <- bind_rows(
  method_overlap(rq2_top_words |> filter(side == "Macho")) |>
    mutate(item_type = "Words", side = "Macho"),
  method_overlap(rq2_top_words |> filter(side == "Nice")) |>
    mutate(item_type = "Words", side = "Nice"),
  method_overlap(rq2_top_traits |> filter(side == "Macho")) |>
    mutate(item_type = "Traits", side = "Macho"),
  method_overlap(rq2_top_traits |> filter(side == "Nice")) |>
    mutate(item_type = "Traits", side = "Nice")
)

kable(
  rq2_top_words,
  caption = "RQ2 top word associates by method",
  digits = 4
)

kable(
  rq2_top_traits,
  caption = "RQ2 top trait associates by method",
  digits = 4
)

kable(
  rq2_method_overlap,
  caption = "RQ2 overlap of top associates across methods",
  digits = 4
)
```

## Q3

**Over time, does the valence of the words or traits most associated with one construct change?**

For each decade and method, I take the top 10 Macho-associated and top 10 Nice-associated items and compute their mean valence on a centered `-4 ... +4` scale by subtracting `5` from Warriner valence.

```{r rq3}
rq3_valence_series <- bind_rows(
  get_top_valence_series(warriner_nudge_method_scores_long, only_traits = FALSE) |>
    mutate(level = "Words"),
  get_top_valence_series(warriner_nudge_method_scores_long, only_traits = TRUE) |>
    mutate(level = "Traits")
) |>
  mutate(
    level = factor(level, levels = c("Words", "Traits")),
    side = factor(side, levels = c("Macho", "Nice"))
  )

rq3_valence_trends <- fit_valence_trends(rq3_valence_series)

rq3_method_agreement <- rq3_valence_series |>
  select(level, side, decade_num, method, mean_valence) |>
  pivot_wider(names_from = method, values_from = mean_valence)

rq3_method_correlations <- map_dfr(
  split(
    rq3_method_agreement,
    interaction(
      rq3_method_agreement$level,
      rq3_method_agreement$side,
      drop = TRUE
    )
  ),
  function(df) {
    map_dfr(combn(c("Pairwise delta", "Anchored", "PLS"), 2, simplify = FALSE), function(pair) {
      tibble(
        level = df$level[1],
        side = df$side[1],
        method_1 = pair[1],
        method_2 = pair[2],
        correlation = cor(df[[pair[1]]], df[[pair[2]]], use = "complete.obs")
      )
    })
  }
)

kable(
  rq3_valence_trends,
  caption = "RQ3 linear trends in mean valence by method",
  digits = 4
)

kable(
  rq3_method_correlations,
  caption = "RQ3 correlation of valence time series across methods",
  digits = 4
)
```

```{r rq3-plot, echo=FALSE, fig.width=11, fig.height=7, fig.align='center'}
ggplot(
  rq3_valence_series,
  aes(x = decade_num, y = mean_valence, color = method)
) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey45") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.7) +
  facet_grid(level ~ side) +
  labs(
    title = "RQ3: valence of top Macho and Nice associates over time",
    x = "Decade",
    y = "Mean centered valence",
    color = "Method"
  ) +
  theme_minimal()
```
