---
title: "Macho-Nice Guy Dichotomy"
author: "Gilad Sarusi"
format: 
  html:
    toc: true
    number-sections: true
    self-contained: true
execute:
  warning: false
  message: false
editor: visual
---

# HistWords repo

For our purposes, I use the same approach chosen by Tessa E. S. Charlesworth and colleagues—namely, the HistWords repository, which is available on [GitHub](https://github.com/williamleif/histwords?tab=readme-ov-file).

### What's going on?

HistWords is an open-source dataset and toolkit that provides word embeddings for different historical periods, trained on Google Books Ngram (and other corpora). Each decade (1800–1990) has its own 300-dimensional vector space representing the meaning of words in that era.

Think of it as a longitudinal semantic dataset—rather than tracking people over time, we track words across decades. Each vector encodes a word’s position in the shared “meaning space” of its time.

#### What can we do with it?

- **Inspect meaning:** find the nearest neighbors of a word in any decade.

- **Compare meanings over time:** compute cosine similarity for the same word across decades to track semantic drift.

- **Study cultural change:** analyze how a word’s associations (e.g., *macho*, *nice*) evolve by decade.

- **Compose phrases:** average vectors (e.g., *macho* + *guy*) to represent multi-word concepts.

- **Align embeddings:** apply sequential Procrustes alignment to make spaces from different decades comparable.

### What are we comparing?

In this project, I adopt an approach similar to Tessa et al. designed to avoid the **alignment problem** inherent in diachronic word embeddings—which makes direct comparison of raw cosine similarities between two embedding spaces (e.g., the 1910s and the 1920s) invalid.

Instead of comparing one decade directly with another, I compare words **within each decade**. In other words, for every decade I ask a simple question: is a given word closer to the *Macho* prototype or to the *Nice Guy* prototype?

To do this, the code builds a vector for each prototype in each decade and then gives every word a **single contrast score**. Positive scores mean the word is closer to the Macho side; negative scores mean it is closer to the Nice side. This lets me describe change across time without pretending that the raw geometry of one decade can be directly compared with the raw geometry of another.

More concretely, the current script does four things:

1.  it builds decade-specific *Macho* and *Nice Guy* vectors in two different ways: a **nudge** version and an **inject** version;
2.  it uses an anchored scoring method to estimate where each word falls on the Macho-versus-Nice contrast within that decade;
3.  it keeps only words that also appear in the Warriner ratings, so later comparisons can be tied to valence; and
4.  in the Attractiveness section, it compares two scoring workflows side by side: the original pairwise cosine approach and the newer anchored approach.

Nevertheless, following Tessa et al.’s guidance, applying **orthogonal Procrustes** rotation can still be useful as a robustness check or for future extensions involving cross-decade word-trajectory analyses.

::: callout-cautio
`The following code is written using the ds4psych: Data Science for Psychology: Natural Language workflow (https://ds4psych.com). Therefore, the toolset used here differs from Tessa’s.`{.smallcaps}
:::



# Start 
```{r loadPackages, echo = FALSE, message=FALSE}
# ----------------------------------------------------
#Packages
library(tidyverse)
library(vroom)       # fast reading of large texts
library(stringr) # string manipulation
library(purrr) # functional programming
library(sweater) # for loading embeddings
library(quanteda.textmodels) # PLS
library(quanteda) # for text analysis
library(embedplyr) # for working with embeddings manipulation
library(knitr) # kable
library(pls) # PLS


```

```{r Datapaths, include=FALSE}

# --- 0) Paths  -------------------

decades <- seq(1800, 1990, by = 10)      
model_names <- setNames(paste0("eng.all_sgns.",decades), decades) # all English books (Google N-grams)


```

Notably, in the 2000s there was a change in N-gram sampling. Therefore, embeddings using the corpus from the 2000s onward are excluded.

```{r}
## Words norm - for valence computation: 


warriner_path <- "wordstim/allwrdnorms_warriner.csv"         
# columns: word, valence, arousal, dominance - valence is the average rating on a 9-point scale from 1 (very negative) to 9 (very positive), obtained from Warriner and colleagues (2013)

# Theoretically we need to cosider that the fact that this was a crowdslabeling procedure taken place in 2013, the chatechristics of each trait in term of valence, arousal and dominance are reflecting the meaning of the word in 2013, which might be different from the meaning of the word in 1800. However, since we are interested in the relative valence of traits (e.g., is "kind" more positive than "dominant"?), and since these relative valence ratings are likely to be more stable over time than absolute ratings, I will proceed with using Warriner's ratings as a proxy for trait valence across all decades. This is a common approach in historical semantics research, although it is important to acknowledge this limitation when interpreting results.

```

As an alternative, we might consider NRC VAD (S. M. Mohammad, 2018), which contains 20,007 words with ratings between 0 and 1 for valence, arousal, and dominance.


### Men Types Word List


```{r MenTypesWordBank}

## List of traits:

traits_path <- "wordstim/traitlist.txt"

# We would latter on create a file storing those words for the code to be more elegant :

nice_guy_words <- c(
  "kind", "gentle", "caring", "sensitive", "considerate", "respectful",
  "supportive", "understanding", "warm", "affectionate",
  "polite", "reliable", "faithful", "honest", "thoughtful",
  "attentive", "humble", "nice"
)
macho_guy_words <- c(
   "dominant", "assertive", "confident", "tough", "strong", "decisive",
  "bold", "powerful", "competitive", "aggressive",  "unemotional",
  "fearless","alpha", "commanding", "ambitious", "macho"
)

```

### Two ways of defining the constructs

To make sure these trait lists are interpreted in a masculine context, I use two slightly different strategies.

The first is to **inject male anchors**. This means I add words such as *man*, *men*, *male*, and *guy* to each list before averaging the embeddings. The goal is to pull the resulting vector toward a masculine semantic context rather than a generic trait context. This is implemented in the *Concepts* chunk.

- **Cons / watch-outs:**\
  *Anchor dominance risk*—male anchors can outweigh trait terms if the lists are small. Moreover, shared anchors in both groups reduce the distance between group centroids (smaller margins).\
  **Monitor:** anchor share in each centroid (≤30–40%), cosine margin between group means, and Top-N stability across decades (overlap %).

The second approach, implemented in the next chunk, is a **masculinity projection nudge**. Here I build a male-versus-female direction in the embedding space and then shift each construct slightly toward the male side before scoring it. This is a softer adjustment than injection: instead of adding anchor words into the list itself, it gently pushes the final vector toward a masculine interpretation.

- **Cons / watch-outs:**\
  *Axis quality*—if the gender sets are unbalanced or drift strongly by decade, the direction can be noisy.\
  *Leakage*—over-weighting (large *w*) can wash out original semantics.\
  **Monitor:** axis stability across decades (cosine between decade-specific *d*<sub>male</sub>), use small *w* (e.g., 0.1–0.3) and run sensitivity analyses; compare results with and without the tweak.

Using both approaches gives me a built-in robustness check. If the results tell the same story under both methods, that makes the interpretation more convincing.

### Masculinity projection nudge {#musculinity-nudge}

To interpret these trait bundles as referring to men rather than to personality in general, I create a female-to-male direction in the semantic space and add part of that direction to the construct vector. In plain language, this step nudges the vector so that later comparisons are more likely to reflect the masculine reading of the traits.

```{r}



# E: numeric matrix (rows = tokens, cols = dims), 

# male_tokens / female_tokens: anchors for the axis (can leave defaults)

 male_anchors <-  c("man","men","male","guy")  #Question: should we add 'boy'? 
 female_anchors <-  c("woman", "women", "female","girl")

 build_ddr <- function(embeddings, words) {
  available_words <- words[words %in% rownames(embeddings)]

  if (length(available_words) == 0) {
    stop("No construct words are available in this decade's embedding space.")
  }

  predict(embeddings, available_words) |>
    average_embedding()
}

 
 build_male_female_axis <- function(embeddings) {
  male_ddr <- build_ddr(embeddings, male_anchors)
  female_ddr <- build_ddr(embeddings, female_anchors)
  male_ddr - female_ddr
}

 

# DDR-based masculinity function:

# w = nudge strength (0 = no nudge, >0 = push toward masculine direction)
 
# default is set to 1 to inteperate results easily and to make sure that the wors be are assigning is regards to word_GUY not that word in general
 
 
masculinity_nudge <- function(ddr_vec, # Or any dim1:dim300 embedding object
                              E, # The embedding space 
                              w = 1) {
 axis <- build_male_female_axis(E)
  axis <- as.numeric(axis)
  
  if (is.matrix(ddr_vec)) {
    
    ddr_vec_nudged <- sweep(
      ddr_vec,
      2,
      w * axis,
      FUN = "+"
    )
    
  } else {
    
    ddr_vec <- as.numeric(ddr_vec)
    ddr_vec_nudged <- ddr_vec + w * axis
    
  }
  
  ddr_vec_nudged

}



```

The second approach, the **Inject approach**, does the same job more directly: it adds male anchor words to each word list before averaging, so the resulting construct vector is explicitly tied to a masculine context.

```{r Concepts, include=FALSE}
# --- 2) Defining concepts (labels) -----------------------------


# Inject them inside the Macho/Nice Construct 
nice_guy_words_injected <- nice_guy_words |>
  c(male_anchors)

macho_guy_words_injected <- macho_guy_words |>
    c(male_anchors)

```

```{r loadData1, include=FALSE}
# --- 3) Loading Data  ----------------


## -------------  Warriner's valence-----------

#(I'm keeping Arousal and Dominance since it might be useful later)

warriner <- read_csv(warriner_path, show_col_types = FALSE) |>
  transmute(word=tolower(Word),
         #Valence
         valence=V.Mean.Sum,
         # Arousal
         arousal=A.Mean.Sum,
         # Dominance
         dominance = D.Mean.Sum,
         id = as.character(row_number())
         ) |>
  
  # dealing with duplicate: averageing across duplicate
  group_by(word) |>
  summarise(
    valence = mean(valence, na.rm = TRUE),
    arousal = mean(arousal, na.rm = TRUE),
    dominance = mean(dominance, na.rm = TRUE),
    id = first(id)
  ) 


## ------------ Lists of traits --------------


trait_list<- read.delim(traits_path, header = FALSE) |>
  # the traits are stored in a column named V1 and I want to use it as a vector 
  pull(V1) 


  # I would fetch only the traits that got crowdslebelling in warriner: 
traits_warriner <- warriner |>
  filter(word %in% trait_list) 
  
```

### ENGall Model

Properties of ENGall_model:

- n of words: 100,000 (same word list across all decades). the authors only computed embeddings for words that were among the top 100,000 most frequent words over all time (for EngAll)
  - n of dimensions: 300.
  - Type of Embedding : SGNS.
  - Context Window : symmetric context window of 4 words.

```{r}
## ----------- ENGall model -----------

if (file.exists("engall/ENGall_model.RDS")) {
  ENGall_model<-readRDS("engall/ENGall_model.RDS")
}else {
  ENGall_model <- lapply(model_names, load_embeddings, save = FALSE) # list of embedding models
  saveRDS(ENGall_model, 
          file="engall/ENGall_model.RDS")
}
length(ENGall_model)
```

The ENGall model provides embeddings for the 100,000 most frequent words **across** all time. This means words coined later in history will be missing in earlier slices, and some words will be absent in particular decade models. The model is stored as a list of 20 elements, one per decade. In some decade data frames, certain words have no embedding.

let's map those words.

```{r}

#### ---- Check unavailable words by decade ----
n_avwords <- vector()
n_unav<- vector()
prop_avail <- vector()
n_total <- vector()

# Creating a list of available word and unavailable words: 

for (i in seq_along(ENGall_model)) {
  # Count available words as those with V1 != 0
  n_avwords[i] <- sum(ENGall_model[[i]][, 1] != 0)
  n_total[i] <- ENGall_model[[i]]|>
      nrow()  
  n_unav[i]  <- n_total[i] - n_avwords[i]
  prop_avail[i] <- n_avwords[i] / n_total[i]
}


avwords_n_decade_df <- data.frame(
  decade      = decades,
  n_avwords   = n_avwords,
  n_unavwords = n_unav,
  prop_avail  = prop_avail
)

```

```{r availWordsTable}
kable(avwords_n_decade_df |>
        #adding comma
        mutate(across(where(is.numeric),scales::comma)),
      digits = 3,
      caption = "Proportion of Words with Embedding")

```

Out of 100,000 words, each decade has on average 36,347 available words.

Notably, the 1990s slice has \~71% of words available. This does **not** mean “29% are post-1990 words.” Rather, 29% of the top-100k (over 1800–1999) simply lack enough occurrences in the 1990s to train vectors—often because they are earlier-era words that faded out, or because they are rare or orthographically different in that decade.

Later, we will pull valence from Warriner’s list (\~14,000 words), which bounds our effective vocabulary to those items. Because not all Warriner words appear in ENGall, the intersection may be smaller.

Another concern is whether some of the words used to define our constructs are missing from the model.

Let's check!

```{r}

### ---- Check if macho/nice words by decade ------- 
# 1) Availability (token x decade)

avail <- map2_df(ENGall_model, decades, ~
  tibble(
    decade    = .y,
    token     = tolower(rownames(.x)),
    available = .x[, 1] != 0
  )
)

#  Define the sets you care about
sets <- list(
  macho = macho_guy_words,
  nice  = nice_guy_words,
  male = male_anchors
)

wanted <- enframe(sets, name = "set", value = "token") |>
  unnest(token) |>
  mutate(token = tolower(token))

# proportion per decade + list of missing words 
macho_nice_words_by_decade_df <- wanted |>
  left_join(avail, by = "token") |>
  mutate(available = tidyr::replace_na(available, FALSE)) |>
  group_by(set, decade) |>
  summarise(
    n_of_words_examined = n(),
    n_available = sum(available),
    prop_avail    = n_available / n_of_words_examined,
    missing     = list(token[!available]),
    .groups = "drop"
  ) |>
  arrange(set, decade)

```

```{r}

kable(macho_nice_words_by_decade_df,
      caption= "Construct's word presence across time")
```

It seems the words *cocky* and *macho* do not appear in ENGall at all. I therefore omit them so that the construct vectors are built only from words that actually exist in the model.

This leaves the Macho and Nice lists with slightly different lengths, which is worth keeping in mind when interpreting the results.

```{r}
length(macho_guy_words)
#[1] 20

length(macho_guy_words_injected)

#[1] 24

length(male_anchors)
# [1] 4

macho_guy_words <- setdiff(macho_guy_words,c("cocky", "macho"))
macho_guy_words_injected <- setdiff(macho_guy_words_injected,c("cocky", "macho"))

# Sanity check
length(macho_guy_words)
# [1] 18

length(macho_guy_words_injected)

# [1] 20

# OK I'm sane
    
    
```

# DDR

I next build the decade-specific semantic representations for *Macho* and *Nice Guy*. In this document I use the term **DDR** simply to mean the vector that summarizes one construct in one decade.

The code creates four such vectors for each decade:

- Macho, using the nudge approach
- Nice Guy, using the nudge approach
- Macho, using the inject approach
- Nice Guy, using the inject approach

## Pairwise

The first scoring method is the *Pairwise Cosine-Delta* method. For every word in the decade-specific embedding space, cosine similarity is computed separately to the Macho DDR and to the Nice DDR. The final score is then defined as:

cosine_delta = cosine(word, Macho DDR) - cosine(word, Nice DDR)

Thus, positive values indicate that a word is more similar to the *Macho than to the Nice*, whereas negative values indicate that a word is more similar to the *Nice than to the Macho*. This method treats the two constructs as two separate reference points and asks whether a word is relatively closer to one than the other- non Dichatomic.

## Anchored Method

The second scoring method is the anchored Macho–Nice axis. Here, the Macho DDR is passed as the positive anchor and the Nice DDR is passed as the negative anchor using

`get_sims(..., method = "anchored").` Rather than computing two independent similarities and subtracting them, the anchored method places each word on a bipolar semantic continuum defined by the Nice pole at one end and the Macho pole at the other. Conceptually, this method asks where each word falls on the semantic axis:

Nice ←———— word ————→ Macho

This scale is ranging form 0 (nice) to 1 (macho). Higher anchored scores indicate a position closer to the Macho pole, whereas lower scores indicate a position closer to the Nice pole. This method therefore treats Macho and Nice as two ends of a single semantic dimension and therefore dichatomic.

## PLS

The third scoring method is the PLS-based Macho–Nice axis. In this approach, a supervised Partial Least Squares model is trained to distinguish Macho seed words from Nice seed words. Macho words are coded as +1, and Nice words are coded as -1. The resulting one-component PLS projection is used as a supervised semantic axis that maximally separates the two word groups. Each word in the decade-specific embedding space is then scored by its cosine similarity to this PLS-derived axis. Positive PLS scores indicate alignment with the Macho-oriented side of the supervised axis, whereas negative scores indicate alignment with the Nice-oriented side.

The three methods therefore answer related but non-identical questions. The pairwise cosine-delta method asks whether a word is more similar to Macho than to Nice. The anchored method asks where a word lies on the bipolar Nice–Macho continuum. The PLS method asks how strongly a word aligns with a supervised axis learned from the Macho versus Nice seed-word distinction. Because these methods produce scores using different geometric procedures, their raw magnitudes should not be interpreted as directly equivalent.

Instead, they should be compared in terms of directional agreement, rank-order similarity, overlap in high-scoring and low-scoring words, and associations with external lexical variables.

```{r helpers-ddr}
make_pls_macho_nice_axis <- function(embeddings,
                                     macho_words,
                                     nice_words,
                                     masculine_nudge = FALSE) {
  
  train_words <- tibble(
    word = c(macho_words, nice_words),
    y = c(
      rep(1, length(macho_words)),
      rep(-1, length(nice_words))
    )
  ) |>
    distinct(word, .keep_all = TRUE) |>
    filter(word %in% rownames(embeddings))

  X <- predict(embeddings, train_words$word)

  if (masculine_nudge) {
    X <- masculinity_nudge(X, E = embeddings)
  }

  X <- as.matrix(X)

  keep <- complete.cases(X) & rowSums(X^2, na.rm = TRUE) > 0

  X <- X[keep, , drop = FALSE]
  y <- train_words$y[keep]

  if (length(y) != nrow(X)) {
    stop("PLS training mismatch: X rows and y labels do not match.")
  }

  if (length(unique(y)) < 2) {
    stop("Only one class remains after removing missing embeddings.")
  }

  pls_data <- as.data.frame(X)
  pls_data$y <- y

  pls_fit <- plsr(
    y ~ .,
    data = pls_data,
    ncomp = 1,
    scale = FALSE,
    validation = "none"
  )

  axis <- as.numeric(pls_fit$projection[, 1])
  scores <- as.numeric(X %*% axis)

  if (mean(scores[y == 1]) < mean(scores[y == -1])) {
    axis <- -axis
  }

  axis / sqrt(sum(axis^2))
}
```

```{r DDRs,warning=FALSE}

# PlS
pls_nudge_axis_by_decade <- vector("list", length(ENGall_model))
pls_inject_axis_by_decade <- vector("list", length(ENGall_model))

Macho_DDR_nudge_by_decade <- vector("list", length(ENGall_model))
Nice_DDR_nudge_by_decade <- vector("list", length(ENGall_model))

Macho_DDR_injected_by_decade <- vector("list", length(ENGall_model))
Nice_DDR_injected_by_decade <- vector("list", length(ENGall_model))

decade_scores <- vector("list", length(ENGall_model))

for (i in seq_along(ENGall_model)) {
  embeddings <- ENGall_model[[i]]
  embeddings <- embeddings[magnitude(embeddings) != 0, ]
  
  # Nudged approach
  Nice_DDR_nudge_by_decade[[i]] <- masculinity_nudge(
    ddr_vec = build_ddr(embeddings, nice_guy_words),
    E = embeddings
  ) 
  
  Macho_DDR_nudge_by_decade[[i]] <- masculinity_nudge(
    ddr_vec = build_ddr(embeddings, macho_guy_words),
    E = embeddings
  ) 
  
  # Inject approach
  Nice_DDR_injected_by_decade[[i]] <- build_ddr(
    embeddings,
    nice_guy_words_injected
  )
  Macho_DDR_injected_by_decade[[i]] <- build_ddr(
    embeddings,
    macho_guy_words_injected
  )
  
  # Method 1: Pairwise Deltas 
  cosine_scores_word_level <- embeddings |>
    get_sims(
      list(
        "macho_cos_nudge" = Macho_DDR_nudge_by_decade[[i]],
        "nice_cos_nudge" = Nice_DDR_nudge_by_decade[[i]],
        "macho_cos_inject" = Macho_DDR_injected_by_decade[[i]],
        "nice_cos_inject" = Nice_DDR_injected_by_decade[[i]]
      ), 
      method = "cosine"
    ) |> 
    mutate(
      cosine_delta_nudge = macho_cos_nudge - nice_cos_nudge,
      cosine_delta_inject = macho_cos_inject - nice_cos_inject
    ) |> 
    select(
      doc_id,
      cosine_delta_nudge,
      cosine_delta_inject
    ) 
  
  # Method 2: Anchored Macho-Nice Axis
  anchored_scores_word_level <- embeddings |> 
    get_sims(
      list(
        "macho_vs_nice_nudge" = list(
          pos = Macho_DDR_nudge_by_decade[[i]],
          neg = Nice_DDR_nudge_by_decade[[i]]
        ), 
        "macho_vs_nice_inject" = list(
          pos = Macho_DDR_injected_by_decade[[i]],
          neg = Nice_DDR_injected_by_decade[[i]]
        ) 
      ), 
      method = "anchored"
    )
  # Method 3: PLS
    pls_nudge_axis_by_decade[[i]] <- make_pls_macho_nice_axis(
    embeddings = embeddings,
    macho_words = macho_guy_words,
    nice_words = nice_guy_words,
    masculine_nudge = TRUE
  )
    
    pls_inject_axis_by_decade[[i]] <- make_pls_macho_nice_axis(
      embeddings = embeddings,
      macho_words = macho_guy_words_injected,
      nice_words = nice_guy_words_injected,
      masculine_nudge = FALSE
    )
    
    pls_scores_word_level <- embeddings |>
      get_sims(
        list(
          "pls_nudge"  = pls_nudge_axis_by_decade[[i]],
          "pls_inject" = pls_inject_axis_by_decade[[i]]
        ),
        method = "cosine"
      ) 
    
  decade_scores[[i]] <- anchored_scores_word_level |>
    left_join(cosine_scores_word_level,
              by = "doc_id")|>
    left_join(
      pls_scores_word_level |>
        select(
          doc_id,
          pls_nudge,
          pls_inject
        ),
      by = "doc_id"
    )
}

# Make it readable
names(Macho_DDR_nudge_by_decade) <- decades
names(Nice_DDR_nudge_by_decade) <- decades
names(Macho_DDR_injected_by_decade) <- decades
names(Nice_DDR_injected_by_decade) <- decades
names(decade_scores) <- decades


```

## Warriner List {#warriner}

Next, I keep **only** the words that also appear in the Warriner norms, because those are the words for which I have valence ratings.

In practice, this means I take the full set of anchored scores from the embedding model and trim it down to the subset of words that can later be interpreted in emotional terms.

```{r WarrinerOnly)}
# Starting with empty 20 long list 

warriner_with_scores <- vector("list", length(decade_scores))

for (i in seq_along(decade_scores)) {

  ds <- decade_scores[[i]]

  # --- standardize the token column name to "word" ---
  
  if ("doc_id" %in% names(ds)) {
    ds2 <- ds |> rename(word= doc_id)
  } else {
    stop(
      "Could not find a word column in decade_scores[[i]]. Check names(decade_scores[[i]])."
      )
  }

  # --- join: keep only Warriner words that exist in this decade vocab ---
  warriner_with_scores[[i]] <-
    warriner |>
    dplyr::inner_join(
      ds2,
      by = "word"
    )
}
names(warriner_with_scores) <- decades

warriner_with_scores


#-------- NOTICE! ------

# Before the comparisons I'll omit the words that were used to create the DDR, namely, macho_guy_word and nice_guy_words, to reduce noise .

omit_words <- unique(c(
  macho_guy_words, nice_guy_words,
  macho_guy_words_injected, nice_guy_words_injected
))

warriner_scores_omit <- lapply(warriner_with_scores, function(df) {
  df |> filter(!word %in% omit_words)
})


```

# RQs

## Q1

**Is there, overall, a difference in association across 200 years between Macho and Nice Guy?**

Here, positive values mean relatively stronger Macho association and negative values mean relatively stronger Nice association. So here we'll see whether the delta, anchored pole loading and pls changes, first overall and then by decade.

In this first section, the tables are used mainly as **descriptive screening summaries**: they show which words, methods, and decades lean toward Macho or Nice on average. The manuscript's formal **longitudinal** inference about change over time is carried out later in the sections that explicitly model decade-to-decade trajectories with tracking-aware weighted `GLS + AR(1)` models.

```{r compare-methods-prepare-long}
method_scores_long <- bind_rows(warriner_scores_omit, .id = "decade") |>
  mutate(
    decade = as.integer(decade),
    decade_num = decade,

    # Anchored scores from get_sims(method = "anchored") are 0-1:
    # 0 = Nice pole, .5 = midpoint, 1 = Macho pole.
    # Center them so 0 = midpoint, positive = Macho, negative = Nice.
    anchored_nudge  = 2 * macho_vs_nice_nudge - 1,
    anchored_inject = 2 * macho_vs_nice_inject - 1,

    
    #Renaming for convience sake 
    pairwise_nudge  = cosine_delta_nudge,
    pairwise_inject = cosine_delta_inject,

    pls_nudge  = pls_nudge,
    pls_inject = pls_inject
  ) |>
  rename()|>
  select(
    word, id, valence, arousal, dominance,
    decade, decade_num,
    anchored_nudge, anchored_inject,
    pairwise_nudge, pairwise_inject,
    pls_nudge, pls_inject
  ) |>
  pivot_longer(
    cols = c(
      anchored_nudge, anchored_inject,
      pairwise_nudge, pairwise_inject,
      pls_nudge, pls_inject
    ),
    names_to = "score_type",
    values_to = "score"
  ) |>
  mutate(
    method = case_when(
      str_detect(score_type, "^anchored") ~ "Anchored",
      str_detect(score_type, "^pairwise") ~ "Pairwise delta",
      str_detect(score_type, "^pls") ~ "PLS"
    ),
    approach = case_when(
      str_detect(score_type, "nudge$") ~ "Nudge",
      str_detect(score_type, "inject$") ~ "Inject"
    ),
    direction = case_when(
      score > 0 ~ "Macho",
      score < 0 ~ "Nice",
      TRUE ~ "Neutral"
    ),
    valence_centered = valence - 5
  ) |>
  filter(is.finite(score))

```

```{r compare-methods-helper}
one_sample_score_summary <- function(x, method) {
  x <- x[is.finite(x)]
  method <- unique(method)

  if (length(method) != 1) {
    stop("one_sample_score_summary() expects exactly one method label per group.")
  }

  null_value <- case_when(
    method == "Anchored" && all(x >= 0 & x <= 1, na.rm = TRUE) ~ 0.5,
    TRUE ~ 0
  )
  
  n <- length(x)
  mean_score <- mean(x)
  sd_score <- if (n > 1) sd(x) else NA_real_
  se <- if (n > 1) sd_score / sqrt(n) else NA_real_
   
  if (n > 1 && is.finite(sd_score) && sd_score > 0) {
    test <- t.test(x, mu = null_value)
     
    tibble(
      n = n,
      mean_score = mean_score,
      sd_score = sd_score,
      se = se,
      null_value = null_value,
      t_value = unname(test$statistic),
      df = unname(test$parameter),
      p_value = test$p.value,
      d_z = (mean_score - null_value) / sd_score,
      ci_low = unname(test$conf.int[1]),
      ci_high = unname(test$conf.int[2])
    )
  } else { 
    tibble(
      n = n,
      mean_score = mean_score,
      sd_score = sd_score,
      se = se,
      null_value = null_value,
      t_value = NA_real_,
      df = NA_real_,
      p_value = NA_real_,
      d_z = NA_real_,
      ci_low = NA_real_,
      ci_high = NA_real_
    )
  }
} 
```

### Word-level analysis

This asks: for each word, is its average score across decades reliably above or below zero?

```{r compare-methods-word-level}
compare_nice_macho_words <- function(data, only_traits = FALSE) {

  if (only_traits) {
    data <- data |>
      filter(word %in% traits_warriner$word)
  }

  data |>
    group_by(method, approach, word) |>
    group_modify(~ one_sample_score_summary(.x$score, method = .y$method)) |>
    ungroup() |>
    left_join(
      data |>
        select(word, valence, arousal, dominance) |>
        distinct(),
      by = "word"
    ) |>
    mutate(
      direction = case_when(
        mean_score > 0 ~ "Macho",
        mean_score < 0 ~ "Nice",
        TRUE ~ "Neutral"
      ),
      p_fdr = p.adjust(p_value, method = "fdr"),
      sig = case_when(
        is.na(p_fdr) ~ "No",
        p_fdr < 0.001 ~ "***",
        p_fdr < 0.01 ~ "**",
        p_fdr < 0.05 ~ "*",
        TRUE ~ "No"
      )
    ) |>
    arrange(method, approach, p_fdr)
}

word_summary_all <- compare_nice_macho_words(
  method_scores_long,
  only_traits = FALSE
)

word_summary_traits <- compare_nice_macho_words(
  method_scores_long,
  only_traits = TRUE
)

word_summary_all |>
  filter(sig != "No") |>
  group_by(method, approach, direction) |>
  arrange(p_fdr, .by_group = TRUE) |>
  slice_head(n = 10) |>
  ungroup() |>
  select(method, approach, direction, word, n, mean_score, d_z, p_value, p_fdr, sig) |>
  kable(
    caption = "Top word-level Macho/Nice scores across decades by method and approach",
    digits = 3
  )

```

### Decade-level analysis

This asks: within each decade, are Warriner words on average shifted toward Macho or Nice?

```{r compare-methods-decade-level}
compare_overall_by_decade <- function(data, only_traits = FALSE) {

  if (only_traits) {
    data <- data |>
      filter(word %in% traits_warriner$word)
  }

  data |>
    group_by(method, approach, decade) |>
    group_modify(~ one_sample_score_summary(.x$score, method = .y$method)) |>
    ungroup() |>
    mutate(
      direction = case_when(
        mean_score > 0 ~ "Macho",
        mean_score < 0 ~ "Nice",
        TRUE ~ "Neutral"
      ),
      sig = case_when(
        is.na(p_value) ~ "No",
        p_value < 0.001 ~ "***",
        p_value < 0.01 ~ "**",
        p_value < 0.05 ~ "*",
        TRUE ~ "No"
      )
    ) |>
    arrange(method, approach, decade)
}

decade_summary_all <- compare_overall_by_decade(
  method_scores_long,
  only_traits = FALSE
)

decade_summary_traits <- compare_overall_by_decade(
  method_scores_long,
  only_traits = TRUE
)

decade_summary_all |>
  select(
    method, approach, decade, n,
    mean_score, d_z, ci_low, ci_high,
    t_value, df, p_value, direction, sig
  ) |>
  kable(
    caption = "Decade-level Macho/Nice shift across Warriner words by method and approach",
    digits = 3
  )


```

### Overall method-level summary

This gives one result per `method × approach` across all word-decade observations.

```{r compare-methods-overall}
overall_summary_all <- method_scores_long |>
  group_by(method, approach) |>
  group_modify(~ one_sample_score_summary(.x$score, method = .y$method)) |>
  ungroup() |>
  mutate(
    direction = case_when(
      mean_score > 0 ~ "Macho",
      mean_score < 0 ~ "Nice",
      TRUE ~ "Neutral"
    ),
    sig = case_when(
      is.na(p_value) ~ "No",
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      TRUE ~ "No"
    )
  )

overall_summary_all |>
  kable(
    caption = "Overall Macho/Nice shift across all Warriner word-decade observations",
    digits = 3
  )
```

## Q2

**Which traits are most consistently pulled toward Macho and Nice Guy across decades?**

To keep this interpretable, I focus here on the rated trait words rather than the full vocabulary. The goal is simple: average each trait's score across decades, keep the significant trait-level results, and then inspect the strongest positive effects (Macho) and strongest negative effects (Nice).

```{r compare-methods-top-traits}
apa_num <- function(x, digits = 2) {
  ifelse(
    is.na(x),
    NA_character_,
    sprintf(paste0("%.", digits, "f"), x)
  )
}

apa_p <- function(p) {
  case_when(
    is.na(p) ~ NA_character_,
    p < 0.001 ~ "< .001",
    TRUE ~ sub("^0\\.", ".", sprintf("%.3f", round(p, 3)))
  )
}

apa_ci <- function(low, high, digits = 2) {
  ifelse(
    is.na(low) | is.na(high),
    NA_character_,
    paste0(
      "[",
      sprintf(paste0("%.", digits, "f"), low),
      ", ",
      sprintf(paste0("%.", digits, "f"), high),
      "]"
    )
  )
}

top_signed_traits <- function(data, top_n = 10) {
  data |>
    filter(sig != "No") |>
    mutate(
      sort_score = if_else(direction == "Macho", d_z, -d_z)
    ) |>
    group_by(method, approach, direction) |>
    slice_max(order_by = sort_score, n = top_n, with_ties = FALSE) |>
    arrange(method, approach, direction, desc(sort_score), .by_group = TRUE) |>
    mutate(rank = row_number()) |>
    ungroup()
}

top_trait_summary <- top_signed_traits(word_summary_traits, top_n = 10)

prepare_trait_table <- function(data, focal_direction) {
  data |>
    filter(direction == focal_direction) |>
    transmute(
      Method = method,
      Approach = approach,
      Rank = rank,
      Trait = word,
      M = apa_num(mean_score, 2),
      `95% CI` = apa_ci(ci_low, ci_high, 2),
      d_z = apa_num(d_z, 2),
      `p (FDR)` = apa_p(p_fdr)
    )
}
```

Before the plot, I keep just the top five traits per method, approach, and direction. If the same trait appears in several method panels, I show it only once within the Macho or Nice row and leave the point visible only in the columns where that trait belongs to the top five. I still use a forest-style plot so the mean trait score and its 95% confidence interval are both visible.

```{r compare-methods-top-traits-plot}
top_trait_plot_data <- top_signed_traits(word_summary_traits, top_n = 5)

trait_row_order <- top_trait_plot_data |>
  group_by(direction, word) |>
  summarise(
    n_panels = n(),
    mean_priority = mean(sort_score, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(direction, desc(n_panels), desc(mean_priority), word) |>
  mutate(direction_word = paste(direction, word, sep = "___"))

top_trait_plot_data <- top_trait_plot_data |>
  left_join(
    trait_row_order |>
      select(direction, word, direction_word),
    by = c("direction", "word")
  ) |>
  mutate(
    direction_word = factor(
      direction_word,
      levels = rev(trait_row_order$direction_word)
    )
  )

ggplot(
  top_trait_plot_data,
  aes(x = mean_score, y = direction_word, color = direction)
) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_segment(
    aes(
      x = ci_low,
      xend = ci_high,
      y = direction_word,
      yend = direction_word
    ),
    linewidth = 0.8,
    show.legend = FALSE
  ) +
  geom_point(size = 2.4, show.legend = FALSE) +
  facet_grid(direction ~ method + approach, scales = "free_y", space = "free_y") +
  scale_y_discrete(labels = function(x) sub("^[^_]+___", "", x)) +
  scale_color_manual(values = c(Macho = "#B35C1E", Nice = "#1C7C7D")) +
  labs(
    title = "Top traits most consistently associated with each construct",
    x = "Mean score across decades with 95% CI",
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text = element_text(face = "bold"),
    panel.grid.major.y = element_blank()
  )
```

The tables below give the same result in a more readable format. I split them into Macho and Nice so each side of the contrast can be scanned on its own.

```{r compare-methods-top-traits-tables}
prepare_trait_table(top_trait_summary, "Macho") |>
  kable(
    caption = "Top Macho-leaning traits across decades",
    align = c("l", "l", "r", "l", "r", "l", "r", "l")
  )

prepare_trait_table(top_trait_summary, "Nice") |>
  kable(
    caption = "Top Nice-leaning traits across decades",
    align = c("l", "l", "r", "l", "r", "l", "r", "l")
  )
```

## Q3

**Over time, does the valence of the traits most associated with each construct change?**

Here I stay with traits only and work directly from the long data. Within each decade, I take the ten most Macho-leaning traits and the ten most Nice-leaning traits, compute their mean centered valence (`valence - 5`), and then track that mean over time. Zero is neutral, positive values are more positive, and negative values are more negative.

```{r compare-methods-top-trait-valence}
select_top_traits_by_decade <- function(data, top_n = 10) {
  data |>
    filter(
      word %in% traits_warriner$word,
      direction != "Neutral"
    ) |>
    mutate(
      rank_score = if_else(direction == "Macho", score, -score)
    ) |>
    group_by(method, approach, decade, decade_num, direction) |>
    slice_max(order_by = rank_score, n = top_n, with_ties = FALSE) |>
    ungroup()
}

top_trait_decade_scores <- select_top_traits_by_decade(
  method_scores_long,
  top_n = 10
)

top_trait_valence_by_decade <- top_trait_decade_scores |>
  group_by(method, approach, decade, decade_num, direction) |>
  summarise(
    n_traits = n(),
    mean_valence_centered = mean(valence_centered, na.rm = TRUE),
    sd_valence_centered = sd(valence_centered, na.rm = TRUE),
    se_valence = if_else(
      n_traits > 1,
      sd_valence_centered / sqrt(n_traits),
      NA_real_
    ),
    t_crit = if_else(
      n_traits > 1,
      qt(0.975, df = n_traits - 1),
      NA_real_
    ),
    ci_low = mean_valence_centered - (t_crit * se_valence),
    ci_high = mean_valence_centered + (t_crit * se_valence),
    mean_score = mean(score, na.rm = TRUE),
    .groups = "drop"
  ) |>
  left_join(
    avwords_n_decade_df |>
      select(decade, n_avwords, prop_avail),
    by = "decade"
  ) |>
  group_by(method, approach, direction) |>
  arrange(decade, .by_group = TRUE) |>
  mutate(
    time_decades = (decade - min(decade)) / 10,
    decade_index = row_number()
  ) |>
  ungroup()
```

The plot shows the decade-by-decade valence trajectory for those top traits under each method and approach. The shaded ribbons are 95% confidence intervals around the decade-specific mean valence.

```{r compare-methods-top-trait-valence-plot}
warriner_valence_plot <- ggplot(
  top_trait_valence_by_decade,
  aes(
    x = decade_num,
    y = mean_valence_centered,
    color = direction,
    fill = direction,
    group = direction
  )
) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_ribbon(
    aes(ymin = ci_low, ymax = ci_high),
    alpha = 0.15,
    color = NA,
    show.legend = FALSE
  ) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.6) +
  facet_grid(approach ~ method) +
  scale_color_manual(values = c(Macho = "#B35C1E", Nice = "#1C7C7D")) +
  scale_fill_manual(values = c(Macho = "#B35C1E", Nice = "#1C7C7D")) +
  labs(
    title = "Mean valence of the top 10 traits within each decade",
    x = "Decade",
    y = "Mean centered valence",
    color = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "top"
  )
```

Before fitting the formal trend model, I first check whether these decade-level valence trajectories show **tracking**. The issue is the same as in the split analysis: if adjacent decades still move together after a basic linear trend is removed, then a standard independent-errors model will understate uncertainty. That would make the valence shift look more precise than it really is.

I therefore fit a simple OLS line within each `method x approach x direction` panel only for diagnostic purposes, and I inspect the lag-1 residual correlation and the largest studentized residual. Positive lag-1 correlations indicate decade-to-decade carryover, whereas very large studentized residuals indicate a potentially influential decade.

```{r compare-methods-top-trait-valence-tracking}
top_trait_valence_tracking_df <- top_trait_valence_by_decade |>
  group_by(method, approach, direction) |>
  group_modify(~ {
    fit <- lm(mean_valence_centered ~ time_decades, data = .x)

    tibble(
      decade = .x$decade,
      residual = resid(fit),
      studentized = rstudent(fit)
    )
  }) |>
  ungroup()

top_trait_valence_tracking_summary <- top_trait_valence_tracking_df |>
  group_by(method, approach, direction) |>
  summarise(
    lag1_residual_correlation = cor(residual[-n()], residual[-1]),
    max_abs_studentized = max(abs(studentized), na.rm = TRUE),
    n_abs_studentized_gt3 = sum(abs(studentized) > 3, na.rm = TRUE),
    .groups = "drop"
  )

kable(
  top_trait_valence_tracking_summary,
  digits = 3,
  caption = "Residual diagnostics for tracking in the top-trait valence trajectories"
)
```

```{r compare-methods-top-trait-valence-tracking-plot}
ggplot(
  top_trait_valence_tracking_df,
  aes(x = decade, y = residual)
) +
  geom_hline(yintercept = 0, linewidth = 0.4) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.4) +
  facet_grid(direction + approach ~ method, scales = "free_y") +
  scale_x_continuous(
    breaks = seq(1800, 1990, by = 20)
  ) +
  labs(
    title = "Residual Tracking Check for Top-Trait Valence Trends",
    x = "Decade",
    y = "OLS residual"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

Once tracking is acknowledged, the more appropriate inferential model is a **weighted GLS with AR(1)**. Here each panel contributes one decade-level mean valence per decade. I use `n_avwords` as the weighting variable so decades with less lexical information contribute less strongly, and I use `AR(1)` so adjacent decades are allowed to remain correlated.

The table below reports the **linear GLS trend** for each panel. The slope is interpreted per decade because `time_decades` is measured in 10-year units. The confidence interval is a Wald interval around that slope estimate.

```{r compare-methods-top-trait-valence-table}
top_trait_valence_trends <- top_trait_valence_by_decade |>
  group_by(method, approach, direction) |>
  group_modify(~ {
    fit <- nlme::gls(
      mean_valence_centered ~ time_decades,
      data = .x,
      correlation = nlme::corAR1(form = ~ decade_index),
      weights = nlme::varFixed(~ I(1 / n_avwords)),
      method = "ML"
    )

    fit_table <- summary(fit)$tTable
    slope_estimate <- unname(fit_table["time_decades", "Value"])
    slope_se <- unname(fit_table["time_decades", "Std.Error"])

    start_row <- .x |>
      arrange(decade) |>
      slice_head(n = 1)

    end_row <- .x |>
      arrange(decade) |>
      slice_tail(n = 1)

    tibble(
      start_decade = start_row$decade,
      start_mean = start_row$mean_valence_centered,
      start_ci_low = start_row$ci_low,
      start_ci_high = start_row$ci_high,
      end_decade = end_row$decade,
      end_mean = end_row$mean_valence_centered,
      end_ci_low = end_row$ci_low,
      end_ci_high = end_row$ci_high,
      change_across_period = end_row$mean_valence_centered - start_row$mean_valence_centered,
      slope_per_decade = slope_estimate,
      slope_ci_low = slope_estimate - qnorm(0.975) * slope_se,
      slope_ci_high = slope_estimate + qnorm(0.975) * slope_se,
      p_value = fit_table["time_decades", "p-value"],
      ar1_phi = unname(
        coef(fit$modelStruct$corStruct, unconstrained = FALSE)
      )
    )
  }) |>
  ungroup()

top_trait_valence_trends |>
  transmute(
    Method = method,
    Approach = approach,
    Direction = direction,
    Start = paste0(
      start_decade,
      ": ",
      apa_num(start_mean, 2),
      " ",
      apa_ci(start_ci_low, start_ci_high, 2)
    ),
    End = paste0(
      end_decade,
      ": ",
      apa_num(end_mean, 2),
      " ",
      apa_ci(end_ci_low, end_ci_high, 2)
    ),
    `Change in M` = apa_num(change_across_period, 2),
    `Slope b` = apa_num(slope_per_decade, 3),
    `95% CI for b` = apa_ci(slope_ci_low, slope_ci_high, 3),
    `AR(1) phi` = apa_num(ar1_phi, 3),
    p = apa_p(p_value)
  ) |>
  kable(
    caption = "Weighted GLS linear change in the valence of the top 10 traits within each decade",
    align = c("l", "l", "l", "l", "l", "r", "r", "l", "r", "l")
  )
```

Because historical change may be curved rather than perfectly linear, I also compare the linear GLS model to a **3 df natural spline GLS model** for each panel. A small likelihood-ratio p-value means the curved model fits better than a single straight line.

```{r compare-methods-top-trait-valence-spline}
top_trait_valence_spline_table <- top_trait_valence_by_decade |>
  group_by(method, approach, direction) |>
  group_modify(~ {
    fit_linear <- nlme::gls(
      mean_valence_centered ~ time_decades,
      data = .x,
      correlation = nlme::corAR1(form = ~ decade_index),
      weights = nlme::varFixed(~ I(1 / n_avwords)),
      method = "ML"
    )

    fit_spline <- nlme::gls(
      mean_valence_centered ~ splines::ns(time_decades, df = 3),
      data = .x,
      correlation = nlme::corAR1(form = ~ decade_index),
      weights = nlme::varFixed(~ I(1 / n_avwords)),
      method = "ML"
    )

    model_comparison <- anova(fit_linear, fit_spline)

    tibble(
      linear_aic = AIC(fit_linear),
      spline_aic = AIC(fit_spline),
      lr_p_value = model_comparison$`p-value`[2],
      better_model = case_when(
        AIC(fit_spline) + 2 < AIC(fit_linear) ~ "Spline",
        AIC(fit_linear) + 2 < AIC(fit_spline) ~ "Linear",
        TRUE ~ "Similar fit"
      )
    )
  }) |>
  ungroup()

top_trait_valence_spline_table |>
  kable(
    digits = 4,
    caption = "Spline sensitivity checks for the top-trait valence trajectories"
  )
```

### Semantic Valence DDR Alternative

The Warriner-based analysis above uses **external valence ratings** that stay fixed across all decades. That is useful, but it also means the valence signal itself does not change historically. As a complementary check, I now build a **decade-specific valence DDR** directly inside each embedding space using straightforward positive-evaluative words such as *good*, *great*, and *wonderful*.

This alternative asks a different question: rather than asking whether the top traits carry high modern valence ratings, it asks where a **positive semantic prototype** falls on the Macho-Nice dimension in each decade. To keep the subsection concise and directly comparable across methods, I use the **nudge-based** Macho-Nice representations and compare the three scoring families only:

- `Pairwise delta`
- `Anchored`
- `PLS`

Because the valence DDR is an average over several seed words, I show uncertainty by **bootstrapping the seed list within each decade**. In each bootstrap sample, the valence words are resampled with replacement, the DDR is rebuilt, and the score is recomputed. The ribbons in the plot are therefore approximate **95% bootstrap intervals** around the valence-construct score for that decade.

```{r valence-ddr-words}
valence_words <- c(
  "good",
  "great",
  "excellent",
  "pleasant",
  "wonderful",
  "lovely",
  "positive",
  "favorable",
  "admirable",
  "delightful"
)

valence_word_coverage <- map_dfr(seq_along(ENGall_model), function(i) {
  embeddings <- ENGall_model[[i]]
  embeddings <- embeddings[magnitude(embeddings) != 0, ]

  tibble(
    decade = decades[i],
    n_valence_words_available = sum(
      valence_words %in% rownames(embeddings)
    )
  )
})

kable(
  valence_word_coverage,
  caption = "Available positive-valence seed words per decade for the semantic valence DDR"
)
```

```{r valence-ddr-analysis}
set.seed(1234)

valence_boot_reps <- 300L

Valence_DDR_by_decade <- vector("list", length(ENGall_model))
valence_ddr_scores_decades <- vector("list", length(ENGall_model))

for (i in seq_along(ENGall_model)) {
  embeddings <- ENGall_model[[i]]
  embeddings <- embeddings[magnitude(embeddings) != 0, ]

  available_valence_words <- valence_words[valence_words %in% rownames(embeddings)]

  Valence_DDR_by_decade[[i]] <- build_ddr(
    embeddings = embeddings,
    words = available_valence_words
  )

  valence_target <- rbind(
    Valence = Valence_DDR_by_decade[[i]]
  )

  central_scores <- tibble(
    method = c("Pairwise delta", "Anchored", "PLS"),
    score = c(
      cos_sim(
        Valence_DDR_by_decade[[i]],
        Macho_DDR_nudge_by_decade[[i]]
      ) - cos_sim(
        Valence_DDR_by_decade[[i]],
        Nice_DDR_nudge_by_decade[[i]]
      ),
      valence_target |>
        get_sims(
          list(
            anchor = list(
              pos = Macho_DDR_nudge_by_decade[[i]],
              neg = Nice_DDR_nudge_by_decade[[i]]
            )
          ),
          method = "anchored"
        ) |>
        as.data.frame() |>
        pull(anchor) |>
        {\(x) x - 0.5}(),
      cos_sim(
        Valence_DDR_by_decade[[i]],
        pls_nudge_axis_by_decade[[i]]
      )
    )
  )

  bootstrap_scores <- map_dfr(seq_len(valence_boot_reps), function(bootstrap_id) {
    sampled_words <- sample(
      available_valence_words,
      size = length(available_valence_words),
      replace = TRUE
    )

    boot_ddr <- build_ddr(
      embeddings = embeddings,
      words = sampled_words
    )

    boot_target <- rbind(
      Valence = boot_ddr
    )

    tibble(
      method = c("Pairwise delta", "Anchored", "PLS"),
      bootstrap_id = bootstrap_id,
      score = c(
        cos_sim(boot_ddr, Macho_DDR_nudge_by_decade[[i]]) -
          cos_sim(boot_ddr, Nice_DDR_nudge_by_decade[[i]]),
        boot_target |>
          get_sims(
            list(
              anchor = list(
                pos = Macho_DDR_nudge_by_decade[[i]],
                neg = Nice_DDR_nudge_by_decade[[i]]
              )
            ),
            method = "anchored"
          ) |>
          as.data.frame() |>
          pull(anchor) |>
          {\(x) x - 0.5}(),
        cos_sim(boot_ddr, pls_nudge_axis_by_decade[[i]])
      )
    )
  })

  valence_ddr_scores_decades[[i]] <- central_scores |>
    left_join(
      bootstrap_scores |>
        group_by(method) |>
        summarise(
          ci_low = quantile(score, 0.025, na.rm = TRUE),
          ci_high = quantile(score, 0.975, na.rm = TRUE),
          .groups = "drop"
        ),
      by = "method"
    ) |>
    mutate(
      decade = decades[i],
      n_valence_words = length(available_valence_words),
      .before = 1
    )
}

names(Valence_DDR_by_decade) <- decades

valence_ddr_plot_df <- bind_rows(valence_ddr_scores_decades) |>
  group_by(method) |>
  arrange(decade, .by_group = TRUE) |>
  mutate(
    time_decades = (decade - min(decade)) / 10
  ) |>
  ungroup()

kable(
  valence_ddr_plot_df |>
    select(decade, method, n_valence_words, score, ci_low, ci_high),
  digits = 3,
  caption = "Decade-specific semantic valence DDR scores across the Macho-Nice methods"
)
```

```{r valence-ddr-plot}
valence_ddr_time_plot <- ggplot(
  valence_ddr_plot_df,
  aes(x = decade, y = score, color = method, fill = method)
) +
  geom_hline(yintercept = 0, linewidth = 0.4) +
  geom_ribbon(
    aes(ymin = ci_low, ymax = ci_high),
    alpha = 0.15,
    color = NA,
    show.legend = FALSE
  ) +
  geom_line(linewidth = 0.8, show.legend = FALSE) +
  geom_point(size = 1.8, show.legend = FALSE) +
  facet_wrap(~ method, scales = "free_y") +
  scale_x_continuous(
    breaks = seq(1800, 1990, by = 20)
  ) +
  scale_color_manual(
    values = c(
      "Pairwise delta" = "#B35C1E",
      "Anchored" = "#1C7C7D",
      "PLS" = "#4C4C4C"
    )
  ) +
  scale_fill_manual(
    values = c(
      "Pairwise delta" = "#B35C1E",
      "Anchored" = "#1C7C7D",
      "PLS" = "#4C4C4C"
    )
  ) +
  labs(
    title = "Semantic Valence DDR Alignment with the Macho-Nice Dimension",
    x = "Decade",
    y = "Valence DDR loading (centered)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

To make the two valence strategies easier to compare, I show them in one vertically stacked figure: the **Warriner-based trait valence trajectory** on top and the **semantic valence DDR trajectory** directly below it.

```{r q3-valence-comparison-plot}
patchwork::wrap_plots(
  warriner_valence_plot,
  valence_ddr_time_plot,
  ncol = 1,
  heights = c(1.15, 0.95)
) +
  patchwork::plot_annotation(
    title = "Comparing Warriner-Based and Semantic-DDR Valence Trajectories"
  )
```
As in the other time-trend sections, I first check for **tracking** before interpreting any formal significance test. Here the concern is the same: if adjacent decades remain similar after a basic linear trend is removed, then a simple independent-errors model would overstate certainty.

```{r valence-ddr-tracking}
valence_ddr_tracking_df <- valence_ddr_plot_df |>
  group_by(method) |>
  group_modify(~ {
    fit <- lm(score ~ time_decades, data = .x)

    tibble(
      decade = .x$decade,
      residual = resid(fit),
      studentized = rstudent(fit)
    )
  }) |>
  ungroup()

valence_ddr_tracking_summary <- valence_ddr_tracking_df |>
  group_by(method) |>
  summarise(
    lag1_residual_correlation = cor(residual[-n()], residual[-1]),
    max_abs_studentized = max(abs(studentized), na.rm = TRUE),
    n_abs_studentized_gt3 = sum(abs(studentized) > 3, na.rm = TRUE),
    .groups = "drop"
  )

kable(
  valence_ddr_tracking_summary,
  digits = 3,
  caption = "Residual diagnostics for tracking in the semantic valence DDR trajectories"
)
```

```{r valence-ddr-tracking-plot}
ggplot(
  valence_ddr_tracking_df,
  aes(x = decade, y = residual)
) +
  geom_hline(yintercept = 0, linewidth = 0.4) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.6) +
  facet_wrap(~ method, scales = "free_y") +
  scale_x_continuous(
    breaks = seq(1800, 1990, by = 20)
  ) +
  labs(
    title = "Residual Tracking Check for Semantic Valence DDR Trends",
    x = "Decade",
    y = "OLS residual"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```
Dependence across decades is not evident in these residual diagnostics, so here I treat the **OLS slope itself** as the formal test of historical change. In other words, for the semantic valence DDR subsection I do **not** move to a GLS correction; instead, I estimate a straightforward linear trend over decades and interpret its slope, confidence interval, and p-value directly.

```{r valence-ddr-ols}
valence_ddr_ols_table <- valence_ddr_plot_df |>
  group_by(method) |>
  group_modify(~ {
    fit <- lm(
      score ~ time_decades,
      data = .x
    )

    fit_summary <- summary(fit)
    fit_ci <- confint(fit)["time_decades", ]
    slope_estimate <- unname(coef(fit)[["time_decades"]])

    start_row <- .x |>
      arrange(decade) |>
      slice_head(n = 1)

    end_row <- .x |>
      arrange(decade) |>
      slice_tail(n = 1)

    tibble(
      start_decade = start_row$decade,
      start_score = start_row$score,
      start_ci_low = start_row$ci_low,
      start_ci_high = start_row$ci_high,
      end_decade = end_row$decade,
      end_score = end_row$score,
      end_ci_low = end_row$ci_low,
      end_ci_high = end_row$ci_high,
      change_across_period = end_row$score - start_row$score,
      slope_per_decade = slope_estimate,
      slope_ci_low = unname(fit_ci[1]),
      slope_ci_high = unname(fit_ci[2]),
      p_value = fit_summary$coefficients["time_decades", "Pr(>|t|)"],
      r_squared = fit_summary$r.squared
    )
  }) |>
  ungroup()

valence_ddr_ols_table |>
  transmute(
    Method = method,
    Start = paste0(
      start_decade,
      ": ",
      apa_num(start_score, 2),
      " ",
      apa_ci(start_ci_low, start_ci_high, 2)
    ),
    End = paste0(
      end_decade,
      ": ",
      apa_num(end_score, 2),
      " ",
      apa_ci(end_ci_low, end_ci_high, 2)
    ),
    `Change in M` = apa_num(change_across_period, 2),
    `Slope b` = apa_num(slope_per_decade, 3),
    `95% CI for b` = apa_ci(slope_ci_low, slope_ci_high, 3),
    `R^2` = apa_num(r_squared, 3),
    p = apa_p(p_value)
  ) |>
  kable(
    caption = "OLS linear change in semantic valence DDR alignment across decades",
    align = c("l", "l", "l", "r", "r", "l", "r", "l")
  )
```

## Q4 -Attractiveness


This section intentionally keeps **both** Attractiveness workflows in the same document.

The older workflow, which came from `main`, asks two separate questions for each decade: "How close is Attractiveness to Macho?" and "How close is Attractiveness to Nice?" The newer workflow, from `anlysis-after-opt`, asks a single contrast question: "Is Attractiveness closer to Macho or to Nice?"

In principle, both approaches should tell a very similar story. Because they currently do not give identical results, I present them side by side and then compare them directly at the end.

### Word list

We begin by specifying a small list of adjectives that directly express attractiveness, such as *attractive*, *beautiful*, *handsome*, and *sexy*. As above, two versions are created:

- Nudge version: uses only the attractiveness descriptors.

- Inject version: includes the attractiveness descriptors plus the male anchor words, ensuring a stronger masculine reference frame in the embedding space.

This parallels the earlier Macho/Nice pipeline for consistency.

```{r attractiveness-words}
attractiveness_words <- c(
  "attractive",
  "handsome",
  "sexy",
  "alluring",
  "charming",
  "appealing"
)

attractiveness_words_injected <- attractiveness_words |>
  c(male_anchors)
```

For each decade, I build an Attractiveness vector using the same two strategies used above for Macho and Nice:

**Inject condition**

1.  Retrieve the embeddings of all attractiveness words + male anchors.
2.  Compute their **average embedding**, yielding the *Attractiveness Injected DDR* for that decade.

**Nudge condition**

1.  Retrieve the embeddings of the attractiveness adjectives.
2.  Average them to obtain the raw *Attractiveness Nudge DDR*.
3.  Apply the **`masculinity_nudge()`** transformation to align the DDR with the nudge paradigm used for the Macho/Nice DDRs.

```{r attractiveness-ddrs}
Attractiveness_DDR_injected_by_decade <- vector("list", length(ENGall_model))
Attractiveness_DDR_nudge_by_decade <- vector("list", length(ENGall_model))

for (i in seq_along(ENGall_model)) {
  embeddings <- ENGall_model[[i]]
  embeddings <- embeddings[magnitude(embeddings) != 0, ]
  
  Attractiveness_DDR_injected_by_decade[[i]] <- build_ddr(
    embeddings = embeddings,
    words = attractiveness_words_injected
  )
  
  Attractiveness_DDR_nudge_by_decade[[i]] <- build_ddr(
    embeddings = embeddings,
    words = attractiveness_words
  ) |>
    masculinity_nudge(E = embeddings)
}

names(Attractiveness_DDR_injected_by_decade) <- decades
names(Attractiveness_DDR_nudge_by_decade) <- decades
```

### DDR analysis

now after we constructed our attractiveness DDR, lets inspect the results of the attractiveness DDR in terms of its loading on the macho/nice axis - across the 3 approach, How does the metric of interest change across years.

```{r Attractiveness-DDR-Analysis}

# raw computation:
cosine_scores_attractives_decades<- vector("list", length(ENGall_model))
pls_scores_attractives_decades<- vector("list", length(ENGall_model))
anchor_scores_attractives_decades<- vector("list", length(ENGall_model))
# cleaned versions: 
cosine_clean_attractives_decades<- vector("list", length(ENGall_model))
pls_clean_attractives_decades<- vector("list", length(ENGall_model))
anchor_clean_attractives_decades<- vector("list", length(ENGall_model))


for (i in seq_along(ENGall_model)) {
  # Method 1: Pairwise Deltas 
  cosine_scores_attractives_decades[[i]] <- rbind(
    macho_nudge_embds = Macho_DDR_nudge_by_decade[[i]],
    nice_nudge_embds  = Nice_DDR_nudge_by_decade[[i]],
    macho_inject_embds = Macho_DDR_injected_by_decade[[i]],
    nice_inject_embds  = Nice_DDR_injected_by_decade[[i]])|> 
    get_sims(
      list(
        "attractiveness_cos_nudge" = Attractiveness_DDR_nudge_by_decade[[i]],
        "attractiveness_cos_inject" = Attractiveness_DDR_injected_by_decade[[i]]
      ),  
      method = "cosine"
    ) |> 
    as.data.frame() |>
    mutate(
      construct = c(
        "macho_nudge_embds",
        "nice_nudge_embds",
        "macho_inject_embds",
        "nice_inject_embds"
      ),
      .before = 1
    )|>
    
    # notice that the data strcuture is less complaing with inject x nudge cells intersect.
    mutate( 
      construct_type = case_when(
        str_detect(construct, "nice")  ~ "nice",
        str_detect(construct, "macho") ~ "macho"
        
      ),
      approach = case_when(
        str_detect(construct, "nudge")  ~ "nudge",
        str_detect(construct, "inject") ~ "inject"
      ),
      cos_sim_attractive = case_when(
        approach == "nudge"  ~ attractiveness_cos_nudge,
        approach == "inject" ~ attractiveness_cos_inject
      )
    ) |>
    select(approach, construct_type, cos_sim_attractive) |>
    pivot_wider(
      names_from = construct_type,
      values_from = cos_sim_attractive
    ) |> 
    mutate( 
      attractive_delta = macho - nice
    )
  # Method 2: Anchored Macho-Nice Axis
  anchor_scores_attractives_decades[[i]] <- rbind(
    Attractivness_inject= Attractiveness_DDR_injected_by_decade[[i]], 
    Attractivness_nudge_ddr = Attractiveness_DDR_nudge_by_decade[[i]]
  ) |>
    get_sims(
      list(
        "macho_vs_nice_inject" = list(
          pos = Macho_DDR_injected_by_decade[[i]],
          neg = Nice_DDR_injected_by_decade[[i]]
        ),
        "macho_vs_nice_nudge" = list(
          pos = Macho_DDR_nudge_by_decade[[i]],
          neg = Nice_DDR_nudge_by_decade[[i]]
        )
      ), 
      method = "anchored"
    ) |> 
    as.data.frame() |>
    mutate(
      construct = c(
        "Attractivness_inject",
        "Attractivness_nudge"
      ),
      .before = 1
    )
  
  # Method 3: PLS Axis 
  
  #reusing the object already created in the previous sections (namely, r DDRs,warning=FALSE)
   
  pls_scores_attractives_decades[[i]] <- rbind(
    Attractivness_nudge_ddr = Attractiveness_DDR_nudge_by_decade[[i]],
    Attractivness_inject= Attractiveness_DDR_injected_by_decade[[i]] 
  ) |>  
    get_sims(
      list(
        "pls_nudge"  = pls_nudge_axis_by_decade[[i]]
      ), 
      method = "cosine"
    ) |> 
    as.data.frame() |>
    mutate(
      construct = c(
        "Attractivness_nudge_ddr"
      ),
      .before = 1
   )
  
  
  
  # Organizing data structure: 
  
  #  each iteration I want two values: delta cosimilarity for each appeaoch 
  cosine_clean_attractives_decades[[i]]<- cosine_scores_attractives_decades[[i]] |>
    mutate( 
      nudge_delta =cosine_scores_attractives_decades[[i]] |>
        filter(approach == "nudge") |>
        pull(attractive_delta),
      inject_delta =cosine_scores_attractives_decades[[i]] |>
        filter(approach == "inject") |>
        pull(attractive_delta)
    ) |>
    select(nudge_delta, inject_delta) |>
    slice(1)
  
  anchor_clean_attractives_decades[[i]]<- anchor_scores_attractives_decades[[i]]|>
    mutate( 
      anchor_inject =anchor_scores_attractives_decades[[i]] |>
        filter(construct == "Attractivness_inject") |>
        pull(macho_vs_nice_inject),
      anchor_nudge =anchor_scores_attractives_decades[[i]] |>
        filter(construct == "Attractivness_nudge") |>
        pull(macho_vs_nice_nudge)
    ) |>
    select(anchor_nudge, anchor_inject) |>
    slice(1) |>
    
    # for easier interpretation would make the sclae beteen -0.5 - 0.5, where -0.5 = nice, 0 = neutral, 0.5 = macho:
    mutate(
      anchor_nudge = anchor_nudge - 0.5,
      anchor_inject = anchor_inject - 0.5
      )

  pls_clean_attractives_decades[[i]] <- pls_scores_attractives_decades[[i]] |>
    mutate( 
      pls_nudge =pls_scores_attractives_decades[[i]] |>
        filter(construct == "Attractivness_nudge_ddr") |>
        pull(pls_nudge)
    ) |>
    select(pls_nudge) |>
    slice(1)
} 
   # summarise and join the cleaned listed with decade as an identifier
cosine_clean_df <- bind_rows(
  cosine_clean_attractives_decades,
  .id = "decade_id"
)|>
  mutate(
    decade = decades[as.integer(decade_id)],
    .before=1
  )

anchor_clean_df <- bind_rows(
  anchor_clean_attractives_decades,
  .id = "decade_id"
)

pls_clean_df <- bind_rows(
  pls_clean_attractives_decades,
  .id = "decade_id"
)

attractiveness_decade_summary <- cosine_clean_df |>
  left_join(anchor_clean_df, by = "decade_id") |>
  left_join(pls_clean_df, by = "decade_id")

```



```{r PlotAttractiveness}

attractiveness_plot_df <- attractiveness_decade_summary |>
  select(
    decade,
    nudge_delta,
    inject_delta,
    anchor_nudge,
    anchor_inject,
    pls_nudge
  ) |>
  pivot_longer(
    cols = -decade,
    names_to = "method",
    values_to = "score"
  ) |>
  mutate(
    approach = case_when(
      str_detect(method, "nudge")  ~ "Nudge",
      str_detect(method, "inject") ~ "Inject"
    ),
    axis_method = case_when(
      str_detect(method, "delta")  ~ "Pairwise cosine delta",
      str_detect(method, "anchor") ~ "Anchored macho–nice axis",
      str_detect(method, "pls")    ~ "PLS axis"
    )
  )


ggplot(
  attractiveness_plot_df,
  aes(
    x = decade,
    y = score,
    linetype = approach,
    group = interaction(axis_method, approach)
  )
) +
  geom_hline(yintercept = 0, linewidth = 0.4) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  facet_wrap(~ axis_method, scales = "free_y") +
  scale_x_continuous(
    breaks = seq(1800, 1990, by = 20)
  ) +
  labs(
    x = "Decade",
    y = "Attractiveness association score",
    linetype = "Embedding construction",
    title = "Association Between Attractiveness and the Macho–Nice Dimension Across Decades"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```
::: callout-note
## What now

Stuff to need to take into consideration to thoroughly continue the analysis

method: 1- Currently the masculinity nudge is set to w=1, meaning that the the nudge is very very storng toward the masculine side. try maybe something more sable. \* should inspect show what happening when varying w

2- the Attractiveness DDR also is constructed using the same "masculine context" embedding both inject and nudge method (which is also set to w=1!) I suspect that the results showed in @warriner that almost all the words loaded greatly more to the macho side of the axis (as shown in *macho_vs_nice* columns) is due to that the attractiveness is also pushed too strongly to the masculinity side.

- should inspect what happend when attractiveness not in the "masculine side" non-inject and non-nudged

future:

3 - Attractiveness as a sexual properties- changing the word bank 4- euclidean distance between the macho and nice guy construct as a support of the existance of a split. \* since only relative distance is possible for comparison across decades, a method is needed to examing the change (distance ) between the constructs. I'm thinking of taking the delta distance of every word from our macho/nice constructs, compute a delta and average across ALL the words and see if this average delta changes across time/

5- what other axaminations would indicate such "stereotype-Split"?
:::

## Q5 - Relationship type 

It does seems that in the literature the only sliant and psychology related findings was that although macho men tend to be percived as more attractive and sexual the appeal prevail only when the relationship type is od casual and non long lasting. When the latter is true, its seems that women are more drawn and prefer the characteristics of the nice guy better than the macho men.

another articles talks about the prefrance of a women in terms of sxual activity and the place in which sexuality takesin their life - *Edward S. Herold & Robin R. Milhausen (1999) Dating preferences of university women: An analysis of the nice guy stereotype, Journal of Sex & Marital Therapy, 25:4, 333-343, DOI: 10.1080/00926239908404010*

To probe that idea directly, I define two new target DDRs. The first is a **long-committed relationship** construct, built from words that emphasize stability, devotion, and pair-bonding. The second is a **short-casual sexual relationship** construct, built from words that emphasize casualness, desire, and sexual gratification.

Because these constructs describe *relationship types* rather than *men themselves*, I keep them as **raw semantic targets**. In other words, I do **not** inject male anchors into the relationship DDRs. Instead, I project the two relationship DDRs onto the **nudge-based** Macho-Nice dimension and ask which relationship type falls closer to the Nice side and which falls closer to the Macho side.

```{r RelationshipTypeWords}
long_committed_relationship_words <- c(
  "committed",
  "faithful",
  "loyal",
  "devoted",
  "marriage",
  "married",
  "partner",
  "romantic",
  "enduring",
  "stable"
)

casual_sexual_relationship_words <- c(
  "casual",
  "sexual",
  "sensual",
  "lust",
  "desire",
  "passionate",
  "promiscuous",
  "physical",
  "pleasure",
  "tempting"
)

relationship_word_coverage <- map_dfr(seq_along(ENGall_model), function(i) {
  embeddings <- ENGall_model[[i]]
  embeddings <- embeddings[magnitude(embeddings) != 0, ]

  tibble(
    decade = decades[i],
    long_committed_available = sum(
      long_committed_relationship_words %in% rownames(embeddings)
    ),
    casual_sexual_available = sum(
      casual_sexual_relationship_words %in% rownames(embeddings)
    )
  )
})

kable(
  relationship_word_coverage,
  caption = "Available words per decade for the relationship-type DDRs"
)
```

```{r RelationshipTypeDDRs}
LongCommitted_DDR_by_decade <- vector("list", length(ENGall_model))
CasualSexual_DDR_by_decade <- vector("list", length(ENGall_model))

for (i in seq_along(ENGall_model)) {
  embeddings <- ENGall_model[[i]]
  embeddings <- embeddings[magnitude(embeddings) != 0, ]

  LongCommitted_DDR_by_decade[[i]] <- build_ddr(
    embeddings = embeddings,
    words = long_committed_relationship_words
  )

  CasualSexual_DDR_by_decade[[i]] <- build_ddr(
    embeddings = embeddings,
    words = casual_sexual_relationship_words
  )
}

names(LongCommitted_DDR_by_decade) <- decades
names(CasualSexual_DDR_by_decade) <- decades
```

### Relationship-type analysis

```{r RelationshipTypeAnalysis}
relationship_scores_decades <- vector("list", length(ENGall_model))

for (i in seq_along(ENGall_model)) {
  relationship_targets <- rbind(
    LongCommitted = LongCommitted_DDR_by_decade[[i]],
    CasualSexual = CasualSexual_DDR_by_decade[[i]]
  )

  delta_scores <- relationship_targets |>
    get_sims(
      list(
        macho = Macho_DDR_nudge_by_decade[[i]],
        nice = Nice_DDR_nudge_by_decade[[i]]
      ),
      method = "cosine"
    ) |>
    as.data.frame()

  anchor_scores <- relationship_targets |>
    get_sims(
      list(
        anchor = list(
          pos = Macho_DDR_nudge_by_decade[[i]],
          neg = Nice_DDR_nudge_by_decade[[i]]
        )
      ),
      method = "anchored"
    ) |>
    as.data.frame()

  pls_scores <- relationship_targets |>
    get_sims(
      list(
        pls = pls_nudge_axis_by_decade[[i]]
      ),
      method = "cosine"
    ) |>
    as.data.frame()

  relationship_scores_decades[[i]] <- tibble(
    construct = c("Long committed", "Casual sexual"),
    decade = decades[i],
    pairwise_delta = delta_scores$macho - delta_scores$nice,
    anchored_axis = anchor_scores$anchor - 0.5,
    pls_axis = pls_scores$pls
  )
}

relationship_plot_df <- bind_rows(relationship_scores_decades) |>
  pivot_longer(
    cols = c(pairwise_delta, anchored_axis, pls_axis),
    names_to = "axis_method",
    values_to = "score"
) |>
  mutate(
    axis_method = case_when(
      axis_method == "pairwise_delta" ~ "Pairwise cosine delta",
      axis_method == "anchored_axis" ~ "Anchored macho-nice axis",
      axis_method == "pls_axis" ~ "PLS axis"
    )
  )

relationship_summary_table <- relationship_plot_df |>
  group_by(axis_method, construct) |>
  summarise(
    mean_score = mean(score, na.rm = TRUE),
    min_score = min(score, na.rm = TRUE),
    max_score = max(score, na.rm = TRUE),
    .groups = "drop"
  )

kable(
  relationship_summary_table,
  digits = 3,
  caption = "Average relationship-type loading on the Macho-Nice dimension across methods"
)
```

```{r RelationshipTypePlot}
ggplot(
  relationship_plot_df,
  aes(
    x = decade,
    y = score,
    color = construct,
    group = construct
  )
) +
  geom_hline(yintercept = 0, linewidth = 0.4) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  facet_wrap(~ axis_method, scales = "free_y") +
  scale_x_continuous(
    breaks = seq(1800, 1990, by = 20)
  ) +
  labs(
    x = "Decade",
    y = "Relationship-type loading",
    color = "Relationship DDR",
    title = "Relationship-Type Alignment with the Macho-Nice Dimension Across Decades"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```

Across the three scoring families, the **long-committed relationship** DDR is expected to sit closer to the **Nice** pole, whereas the **casual-sexual relationship** DDR should sit relatively closer to the **Macho** pole. In practice, the key question is not whether the casual-sexual DDR always becomes fully Macho-coded, but whether it is **consistently less Nice-coded** than the long-committed DDR. If that pattern holds, it would mirror the literature's claim that the Macho stereotype is more compatible with short-term sexual appeal, whereas the Nice stereotype is more compatible with long-term commitment.



## Q6 - Split Growth Across Decades


###Checking for Tracking Before Trend Modeling

Before fitting any formal time-trend model, I first check whether the decade series shows signs of **tracking**, meaning that adjacent decades carry similar residual error. This matters because a simple linear model assumes that once the overall trend is fitted, the remaining errors are independent. In diachronic semantic data that assumption is often unrealistic: the meaning of masculinity in 1900 is not generated from scratch, but evolves gradually from 1890 and 1910.

If residuals track each other over time, then an ordinary linear model becomes overconfident. In practical terms, that means:

- standard errors can be too small;
- p-values can be too optimistic;
- confidence intervals can be too narrow.

To run that diagnostic, I first define the split metrics themselves. I use two families of split indicators:

1.  **prototype distance**: how far apart the Macho and Nice Guy DDRs are in a given decade;
2.  **lexicon-wide polarization**: how strongly the whole decade lexicon is pulled toward one pole versus the other.

Normalization matters here because a DDR is an **average vector**, and average vectors are **not** unit length. In cosine-based measures this is not a problem, because cosine is scale-invariant: normalizing a vector does **not** change its cosine relationship to another vector. In Euclidean distance, however, vector length does matter. A larger **raw** Euclidean distance can reflect either:

1.  a stronger directional split between Macho and Nice;  
2.  or simply longer DDR vectors in that decade.

For that reason, I report both **raw Euclidean distance** and **normalized Euclidean distance**:

- If **both** raw and normalized distance grow, the split is strengthening in a robust geometric sense.
- If **raw** distance grows but **normalized** distance does not, the constructs are becoming stronger in magnitude, but not necessarily more directionally distinct.
- If **normalized** distance grows while raw distance stays flat, the two constructs are becoming more directionally opposed even without larger vector norms.

I also report the average magnitude of the Macho-versus-Nice contrast across the full decade lexicon.

```{r SplitMetrics}
split_metrics_decades <- vector("list", length(ENGall_model))

for (i in seq_along(ENGall_model)) {
  embeddings <- ENGall_model[[i]]
  embeddings <- embeddings[magnitude(embeddings) != 0, ]

  macho_ddr_raw <- as.numeric(Macho_DDR_nudge_by_decade[[i]])
  nice_ddr_raw <- as.numeric(Nice_DDR_nudge_by_decade[[i]])

  macho_ddr_norm <- normalize(macho_ddr_raw)
  nice_ddr_norm <- normalize(nice_ddr_raw)

  delta_scores <- embeddings |>
    get_sims(
      list(
        macho = macho_ddr_norm,
        nice = nice_ddr_norm
      ),
      method = "cosine"
    ) |>
    as.data.frame() |>
    transmute(delta = macho - nice)

  split_metrics_decades[[i]] <- tibble(
    decade = decades[i],
    macho_norm = magnitude(macho_ddr_raw),
    nice_norm = magnitude(nice_ddr_raw),
    cosine_similarity = cos_sim(macho_ddr_raw, nice_ddr_raw),
    cosine_distance = 1 - cos_sim(macho_ddr_raw, nice_ddr_raw),
    raw_euclidean_distance = euc_dist(macho_ddr_raw, nice_ddr_raw),
    normalized_euclidean_distance = euc_dist(macho_ddr_norm, nice_ddr_norm),
    mean_abs_delta = mean(abs(delta_scores$delta), na.rm = TRUE),
    sd_delta = sd(delta_scores$delta, na.rm = TRUE)
  )
}

split_metrics_df <- bind_rows(split_metrics_decades)

kable(
  split_metrics_df,
  digits = 3,
  caption = "Distance and polarization between the Macho and Nice Guy constructs across decades"
)
```

To diagnose tracking, I intentionally fit a **simple ordinary least squares (OLS)** line to each split metric. This is *not* the final model. It is a diagnostic baseline whose residuals let us ask a simple question: after removing a basic linear time trend, do adjacent decades still move together?

I first prepare the time index and also join the decade-level embedding availability information. The availability values are not needed for the residual diagnostic itself, but they will be used in the weighted models that follow.

```{r SplitModelPrep}
split_model_df <- split_metrics_df |>
  left_join(
    avwords_n_decade_df |>
      select(decade, n_avwords, prop_avail),
    by = "decade"
  ) |>
  mutate(
    time_decades = (decade - min(decade)) / 10,
    decade_index = row_number()
  )
```

In the next chunk, I extract two diagnostic quantities from the simple OLS fits:

- the **raw residuals**, which show whether adjacent decades tend to stay above or below the fitted line together;
- the **studentized residuals**, which scale each residual by its estimated uncertainty. Values larger than about `|3|` are often treated as unusually extreme.

If residuals show long runs of the same sign, or if the lag-1 residual correlation is high, that is evidence of tracking and therefore evidence that a simple independent-errors model is not appropriate.

```{r SplitTrackingDiagnostics}
split_tracking_diagnostics_df <- c(
  "cosine_distance",
  "raw_euclidean_distance",
  "normalized_euclidean_distance",
  "mean_abs_delta",
  "sd_delta"
) |>
  map_dfr(function(metric) {
    fit <- lm(reformulate("time_decades", response = metric), data = split_model_df)

    tibble(
      decade = split_model_df$decade,
      metric = metric,
      fitted = fitted(fit),
      residual = resid(fit),
      studentized = rstudent(fit)
    )
  }) |>
  mutate(
    metric = case_when(
      metric == "cosine_distance" ~ "Prototype cosine distance",
      metric == "raw_euclidean_distance" ~ "Prototype Euclidean distance (raw)",
      metric == "normalized_euclidean_distance" ~ "Prototype Euclidean distance (normalized)",
      metric == "mean_abs_delta" ~ "Mean absolute delta across lexicon",
      metric == "sd_delta" ~ "SD of delta across lexicon"
    )
  )
```

The next table summarizes the most important diagnostic signals. The **lag-1 residual correlation** tells us how similar each residual is to the next decade's residual. Values clearly above zero indicate temporal carryover. The **maximum absolute studentized residual** and the number of residuals larger than `|3|` help identify whether isolated extreme decades are also contributing to the problem.

```{r SplitTrackingSummary}
split_tracking_summary_table <- split_tracking_diagnostics_df |>
  group_by(metric) |>
  summarise(
    lag1_residual_correlation = cor(residual[-n()], residual[-1]),
    max_abs_studentized = max(abs(studentized), na.rm = TRUE),
    n_abs_studentized_gt3 = sum(abs(studentized) > 3, na.rm = TRUE),
    .groups = "drop"
  )

kable(
  split_tracking_summary_table,
  digits = 3,
  caption = "Residual diagnostics for detecting tracking in the decade series"
)
```

I looks like the lag-1 is moderate-high (.35-.42) across all split metrics, indicating that adjacent decades are not independent. The maximum absolute studentized residual is largest for raw Euclidean distance, mean absolute delta, and SD of delta, suggesting that the earliest decade is particularly influential for those measures.

#### Plot
I now plot the raw residuals against time. If the residuals are independent, these panels should look patternless. If nearby decades cluster together above or below zero, that is visual evidence of tracking.

```{r SplitTrackingResidualPlot}
ggplot(
  split_tracking_diagnostics_df,
  aes(x = decade, y = residual)
) +
  geom_hline(yintercept = 0, linewidth = 0.4) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  facet_wrap(~ metric, scales = "free_y") +
  scale_x_continuous(
    breaks = seq(1800, 1990, by = 20)
  ) +
  labs(
    x = "Decade",
    y = "OLS residual",
    title = "Residual Tracking Check Across Decades"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank()
  )
```

Finally, I inspect the studentized residuals. These plots help separate two issues:

1.  **tracking**, where many adjacent residuals move together;
2.  **single influential decades**, where one or two decades stand out sharply from the rest.

##### Studentized Resudual
```{r SplitTrackingStudentizedPlot}
ggplot(
  split_tracking_diagnostics_df,
  aes(x = decade, y = studentized)
) +
  geom_hline(yintercept = 0, linewidth = 0.4) +
  geom_hline(yintercept = c(-3, 3), linetype = "dashed", linewidth = 0.4) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  facet_wrap(~ metric, scales = "free_y") +
  scale_x_continuous(
    breaks = seq(1800, 1990, by = 20)
  ) +
  labs(
    x = "Decade",
    y = "Studentized residual",
    title = "Studentized Residual Check Across Decades"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank()
  )
```

Residual diagnostics showed moderate positive tracking across all split metrics (lag-1 residual correlation = .35-.42), indicating that adjacent decades were not statistically independent. Studentized residuals further revealed that the earliest decade was highly influential for magnitude-sensitive measures, particularly raw Euclidean distance, mean absolute delta, and SD of delta, whereas cosine and normalized Euclidean distance remained below the conventional outlier threshold. 

This pattern suggests that early-decade instability affects raw magnitude-based measures more strongly than direction-based measures, supporting the use of weighted GLS models with AR(1) errors and robustness checks excluding unstable early decades.



### Analysis 
The next step is to model whether these split metrics show a **systematic trend over time** once tracking is handled explicitly.

At the same time, not all decades are equally informative. The ENGall availability table above shows that some decades have more words with usable embeddings than others. If a decade has fewer usable vectors, then its split estimate is likely to be less precise. To reflect that, I add a **Weighted Least Squares (WLS)** component to the model.

Because the total ENGall vocabulary is fixed across decades, `n_avwords` and `prop_avail` contain the same ranking information. I therefore use `n_avwords` directly because it is more concrete to interpret: it is the number of words with usable embeddings in that decade.

The combined model is therefore a **weighted GLS with AR(1)**. In plain language:

- `GLS` estimates the trend line while allowing the residuals to have a structured dependence.
- `AR(1)` means each decade is allowed to be statistically correlated with the immediately adjacent decade.
- `WLS` gives less influence to decades with fewer usable embeddings, and more influence to decades with richer lexical coverage.
- `phi` is the autocorrelation parameter: values closer to `1` mean stronger decade-to-decade carryover.

In `nlme::gls()`, the weighting is expressed as a **variance function** rather than as simple case weights. I specify `weights = varFixed(~ I(1 / n_avwords))`, which means decades with fewer available words are treated as having larger residual variance and therefore contribute less to the fitted trend.

I now fit the **linear weighted GLS model** for each split metric. This tells us whether the split tends to increase or decrease over time once both autocorrelation and uneven decade-level information are taken into account.

```{r SplitGLSLinear}
split_gls_linear_table <- c(
  "cosine_distance",
  "raw_euclidean_distance",
  "normalized_euclidean_distance",
  "mean_abs_delta",
  "sd_delta"
) |>
  map_dfr(function(metric) {
    fit <- nlme::gls( # this is a normal generalized least squares model, but with AR(1) correlation structure and weights for the number of available words
      reformulate("time_decades", response = metric),
      data = split_model_df,
      correlation = nlme::corAR1(form = ~ decade_index),
      weights = nlme::varFixed(~ I(1 / n_avwords)),
      method = "ML"
    )

    tibble(
      metric = metric,
      slope_per_decade = unname(coef(fit)[["time_decades"]]),
      p_value = summary(fit)$tTable["time_decades", "p-value"],
      ar1_phi = unname(
        coef(fit$modelStruct$corStruct, unconstrained = FALSE)
      ),
      aic = AIC(fit)
    )
  })

split_gls_linear_table <- split_gls_linear_table |>
  mutate(
    metric = case_when(
      metric == "cosine_distance" ~ "Prototype cosine distance",
      metric == "raw_euclidean_distance" ~ "Prototype Euclidean distance (raw)",
      metric == "normalized_euclidean_distance" ~ "Prototype Euclidean distance (normalized)",
      metric == "mean_abs_delta" ~ "Mean absolute delta across lexicon",
      metric == "sd_delta" ~ "SD of delta across lexicon"
    )
  )

kable(
  split_gls_linear_table,
  digits = 4,
  caption = "Weighted GLS linear trend tests for whether the Macho-Nice split grows across decades"
)
```

Language change over two centuries may not be perfectly linear, so I also fit a **natural spline** version of the same weighted GLS model. A natural spline allows the trend to bend gradually rather than forcing one straight line through the whole period. This is a useful sensitivity check because cultural change often unfolds in phases rather than at a constant rate.

To compare the linear and spline models fairly, both are fit with **maximum likelihood (`ML`)**. That matters because when the fixed-effects structure changes, `ML` is the standard basis for comparing model fit. I then compare the two models using:

- `AIC`: lower values indicate better fit with a penalty for extra complexity;
- `L.Ratio p-value`: whether the spline improves fit enough over the linear model to justify the added flexibility.

```{r SplitGLSSpline}
split_gls_spline_table <- c(
  "cosine_distance",
  "raw_euclidean_distance",
  "normalized_euclidean_distance",
  "mean_abs_delta",
  "sd_delta"
) |>
  map_dfr(function(metric) {
    fit_linear <- nlme::gls( #This is a normal generalized least squares model, but with AR(1) correlation structure and weights for the number of available words
      reformulate("time_decades", response = metric),
      data = split_model_df,
      correlation = nlme::corAR1(form = ~ decade_index),
      weights = nlme::varFixed(~ I(1 / n_avwords)),
      method = "ML"
    )

    fit_spline <- nlme::gls(# this is a model that allows for a non-linear relationship between time and the split metric, using natural splines with 3 degrees of freedom (curves)
      reformulate("splines::ns(time_decades, df = 3)", response = metric),
      data = split_model_df,
      correlation = nlme::corAR1(form = ~ decade_index),
      weights = nlme::varFixed(~ I(1 / n_avwords)),
      method = "ML"
    )

    model_comparison <- anova(fit_linear, fit_spline)

    tibble(
      metric = metric,
      linear_aic = AIC(fit_linear),
      spline_aic = AIC(fit_spline),
      lr_p_value = model_comparison$`p-value`[2],
      better_model = case_when(
        AIC(fit_spline) + 2 < AIC(fit_linear) ~ "Spline",
        AIC(fit_linear) + 2 < AIC(fit_spline) ~ "Linear",
        TRUE ~ "Similar fit"
      )
    )
  })

split_gls_spline_table <- split_gls_spline_table |>
  mutate(
    #Renaming 
    metric = case_when(
      metric == "cosine_distance" ~ "Prototype cosine distance",
      metric == "raw_euclidean_distance" ~ "Prototype Euclidean distance (raw)",
      metric == "normalized_euclidean_distance" ~ "Prototype Euclidean distance (normalized)",
      metric == "mean_abs_delta" ~ "Mean absolute delta across lexicon",
      metric == "sd_delta" ~ "SD of delta across lexicon"
    )
  )

kable(
  split_gls_spline_table,
  digits = 4,
  caption = "Weighted GLS spline sensitivity checks for nonlinear change in the Macho-Nice split"
)
```

Before looking at the autocorrelation-aware models, it is still useful to inspect the raw trajectories directly. The next plot shows the observed split metrics by decade, without imposing a fitted trend line.

```{r SplitPlot}
split_plot_df <- split_metrics_df |>
  select(
    decade,
    cosine_distance,
    raw_euclidean_distance,
    normalized_euclidean_distance,
    mean_abs_delta,
    sd_delta
  ) |>
  pivot_longer(
    cols = -decade,
    names_to = "metric",
    values_to = "value"
  ) |>
  mutate(
    metric = case_when(
      metric == "cosine_distance" ~ "Prototype cosine distance",
      metric == "raw_euclidean_distance" ~ "Prototype Euclidean distance (raw)",
      metric == "normalized_euclidean_distance" ~ "Prototype Euclidean distance (normalized)",
      metric == "mean_abs_delta" ~ "Mean absolute delta across lexicon",
      metric == "sd_delta" ~ "SD of delta across lexicon"
    )
  )

ggplot(
  split_plot_df,
  aes(x = decade, y = value)
) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  facet_wrap(~ metric, scales = "free_y") +
  scale_x_continuous(
    breaks = seq(1800, 1990, by = 20)
  ) +
  labs(
    x = "Decade",
    y = "Split metric",
    title = "Growth of the Macho-Nice Split Across Decades"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank()
  )
```

### Robustness Check: Excluding Unstable Early Decades

The first two decades appear visually extreme across several panels, especially in the lexicon-wide delta metrics. To check whether these early values are driving the results, I rerun the same split analysis after excluding **1800** and **1810** as potentially unstable decades.

The logic here is straightforward: if the substantive pattern is real, it should survive after removing the decades that look unusually influential. If the significance or direction changes sharply, that suggests the full-sample result may depend heavily on those early observations.

I first rebuild the time index on the reduced sample. This is important because the AR(1) model should treat the filtered series as a new sequence of equally spaced decade observations. I keep the same availability-based weighting logic here as well, so the reduced-sample robustness check remains directly comparable to the full-sample model.

```{r SplitRobustnessPrep}
unstable_decades <- c(1800, 1810)

split_metrics_stable_df <- split_metrics_df |>
  left_join(
    avwords_n_decade_df |>
      select(decade, n_avwords, prop_avail),
    by = "decade"
  ) |>
  filter(!decade %in% unstable_decades) |>
  mutate(
    time_decades = (decade - min(decade)) / 10,
    decade_index = row_number()
  )
```

Before refitting the reduced-sample trend models, I again check whether tracking remains in the filtered decade series. This matters because dropping influential early decades can reduce outlier pressure without fully removing temporal dependence. If adjacent residuals still move together, the reduced-sample inference should still use the same `GLS + AR(1)` structure rather than reverting to an independent-errors model.

```{r SplitRobustnessTracking}
split_tracking_stable_summary_table <- c(
  "cosine_distance",
  "raw_euclidean_distance",
  "normalized_euclidean_distance",
  "mean_abs_delta",
  "sd_delta"
) |>
  map_dfr(function(metric) {
    fit <- lm(
      reformulate("time_decades", response = metric),
      data = split_metrics_stable_df
    )

    tibble(
      metric = metric,
      lag1_residual_correlation = cor(resid(fit)[-length(resid(fit))], resid(fit)[-1]),
      max_abs_studentized = max(abs(rstudent(fit)), na.rm = TRUE),
      n_abs_studentized_gt3 = sum(abs(rstudent(fit)) > 3, na.rm = TRUE)
    )
  }) |>
  mutate(
    metric = case_when(
      metric == "cosine_distance" ~ "Prototype cosine distance",
      metric == "raw_euclidean_distance" ~ "Prototype Euclidean distance (raw)",
      metric == "normalized_euclidean_distance" ~ "Prototype Euclidean distance (normalized)",
      metric == "mean_abs_delta" ~ "Mean absolute delta across lexicon",
      metric == "sd_delta" ~ "SD of delta across lexicon"
    )
  )

kable(
  split_tracking_stable_summary_table,
  digits = 3,
  caption = "Residual diagnostics for tracking after excluding 1800 and 1810"
)
```

I then refit the same **weighted linear GLS + AR(1)** models on the reduced sample.

```{r SplitRobustnessLinear}
split_gls_linear_stable_table <- c(
  "cosine_distance",
  "raw_euclidean_distance",
  "normalized_euclidean_distance",
  "mean_abs_delta",
  "sd_delta"
) |>
  map_dfr(function(metric) {
    fit <- nlme::gls(
      reformulate("time_decades", response = metric),
      data = split_metrics_stable_df,
      correlation = nlme::corAR1(form = ~ decade_index),
      weights = nlme::varFixed(~ I(1 / n_avwords)),
      method = "ML"
    )

    tibble(
      metric = metric,
      slope_per_decade = unname(coef(fit)[["time_decades"]]),
      p_value = summary(fit)$tTable["time_decades", "p-value"],
      ar1_phi = unname(
        coef(fit$modelStruct$corStruct, unconstrained = FALSE)
      ),
      aic = AIC(fit)
    )
  })

split_gls_linear_stable_table <- split_gls_linear_stable_table |>
  mutate(
    metric = case_when(
      metric == "cosine_distance" ~ "Prototype cosine distance",
      metric == "raw_euclidean_distance" ~ "Prototype Euclidean distance (raw)",
      metric == "normalized_euclidean_distance" ~ "Prototype Euclidean distance (normalized)",
      metric == "mean_abs_delta" ~ "Mean absolute delta across lexicon",
      metric == "sd_delta" ~ "SD of delta across lexicon"
    )
  )

kable(
  split_gls_linear_stable_table,
  digits = 4,
  caption = "Weighted GLS linear trend tests for the Macho-Nice split after excluding 1800 and 1810"
)
```

Finally, I repeat the **weighted spline sensitivity check** on the reduced sample. This tells us whether any nonlinear pattern still appears once the earliest decades are removed.

```{r SplitRobustnessSpline}
split_gls_spline_stable_table <- c(
  "cosine_distance",
  "raw_euclidean_distance",
  "normalized_euclidean_distance",
  "mean_abs_delta",
  "sd_delta"
) |>
  map_dfr(function(metric) {
    fit_linear <- nlme::gls(
      reformulate("time_decades", response = metric),
      data = split_metrics_stable_df,
      correlation = nlme::corAR1(form = ~ decade_index),
      weights = nlme::varFixed(~ I(1 / n_avwords)),
      method = "ML"
    )

    fit_spline <- nlme::gls(
      reformulate("splines::ns(time_decades, df = 3)", response = metric),
      data = split_metrics_stable_df,
      correlation = nlme::corAR1(form = ~ decade_index),
      weights = nlme::varFixed(~ I(1 / n_avwords)),
      method = "ML"
    )

    model_comparison <- anova(fit_linear, fit_spline)

    tibble(
      metric = metric,
      linear_aic = AIC(fit_linear),
      spline_aic = AIC(fit_spline),
      lr_p_value = model_comparison$`p-value`[2],
      better_model = case_when(
        AIC(fit_spline) + 2 < AIC(fit_linear) ~ "Spline",
        AIC(fit_linear) + 2 < AIC(fit_spline) ~ "Linear",
        TRUE ~ "Similar fit"
      )
    )
  })

split_gls_spline_stable_table <- split_gls_spline_stable_table |>
  mutate(
    metric = case_when(
      metric == "cosine_distance" ~ "Prototype cosine distance",
      metric == "raw_euclidean_distance" ~ "Prototype Euclidean distance (raw)",
      metric == "normalized_euclidean_distance" ~ "Prototype Euclidean distance (normalized)",
      metric == "mean_abs_delta" ~ "Mean absolute delta across lexicon",
      metric == "sd_delta" ~ "SD of delta across lexicon"
    )
  )

kable(
  split_gls_spline_stable_table,
  digits = 4,
  caption = "Weighted GLS spline sensitivity checks for the Macho-Nice split after excluding 1800 and 1810"
)
```

Instead of showing only the observed reduced-sample trajectories, I now plot the **fitted models themselves**. For each metric, I refit the full-sample and reduced-sample models, select the better-fitting specification between the **linear GLS + AR(1)** and **spline GLS + AR(1)** versions using the same AIC rule as above, and then plot the model-implied trajectory over the shared **1820--1990** period.

The shaded ribbons are approximate **95% confidence intervals for the fitted mean trajectory**, not prediction intervals for a single decade. This is the most direct visual robustness check: if the reduced-sample fit stays close to the full-sample fit, then the substantive conclusion is not being driven mainly by the excluded early decades.

```{r SplitRobustnessPlot}
metric_labels <- c(
  cosine_distance = "Prototype cosine distance",
  raw_euclidean_distance = "Prototype Euclidean distance (raw)",
  normalized_euclidean_distance = "Prototype Euclidean distance (normalized)",
  mean_abs_delta = "Mean absolute delta across lexicon",
  sd_delta = "SD of delta across lexicon"
)

shared_decades <- setdiff(decades, unstable_decades)

split_robustness_model_plot_df <- names(metric_labels) |>
  map_dfr(function(metric) {
    tibble(
      sample = c("Full sample", "Excluding 1800 and 1810"),
      model_data = list(split_model_df, split_metrics_stable_df)
    ) |>
      mutate(
        plot_df = map2(sample, model_data, function(sample, model_data) {
          fit_linear <- nlme::gls(
            reformulate("time_decades", response = metric),
            data = model_data,
            correlation = nlme::corAR1(form = ~ decade_index),
            weights = nlme::varFixed(~ I(1 / n_avwords)),
            method = "ML"
          )

          fit_spline <- nlme::gls(
            reformulate("splines::ns(time_decades, df = 3)", response = metric),
            data = model_data,
            correlation = nlme::corAR1(form = ~ decade_index),
            weights = nlme::varFixed(~ I(1 / n_avwords)),
            method = "ML"
          )

          selected_fit <- if (AIC(fit_spline) + 2 < AIC(fit_linear)) {
            fit_spline
          } else {
            fit_linear
          }

          selected_model <- if (AIC(fit_spline) + 2 < AIC(fit_linear)) {
            "Spline GLS + AR(1)"
          } else {
            "Linear GLS + AR(1)"
          }

          newdata <- tibble(
            decade = shared_decades,
            time_decades = (shared_decades - min(model_data$decade)) / 10
          )

          design_matrix <- model.matrix(
            delete.response(terms(selected_fit)),
            data = newdata
          )

          fit_value <- as.numeric(design_matrix %*% coef(selected_fit))
          fit_se <- sqrt(
            diag(
              design_matrix %*%
                vcov(selected_fit) %*%
                t(design_matrix)
            )
          )

          tibble(
            decade = shared_decades,
            metric = metric_labels[[metric]],
            sample = sample,
            selected_model = selected_model,
            fitted = fit_value,
            conf_low = fit_value - qnorm(0.975) * fit_se,
            conf_high = fit_value + qnorm(0.975) * fit_se
          )
        })
      ) |>
      select(plot_df) |>
      unnest(plot_df)
  })

ggplot(
  split_robustness_model_plot_df,
  aes(x = decade, y = fitted, color = sample, fill = sample)
) +
  geom_ribbon(
    aes(ymin = conf_low, ymax = conf_high),
    alpha = 0.16,
    linewidth = 0,
    color = NA
  ) +
  geom_line(
    aes(linetype = selected_model),
    linewidth = 0.9
  ) +
  facet_wrap(~ metric, scales = "free_y") +
  scale_x_continuous(
    breaks = seq(1820, 1990, by = 20)
  ) +
  scale_color_manual(
    values = c(
      "Full sample" = "black",
      "Excluding 1800 and 1810" = "firebrick"
    )
  ) +
  scale_fill_manual(
    values = c(
      "Full sample" = "grey55",
      "Excluding 1800 and 1810" = "firebrick"
    )
  ) +
  labs(
    x = "Decade",
    y = "Model-implied split metric",
    color = "Sample",
    fill = "Sample",
    linetype = "Selected model",
    title = "Model-Based Robustness Check for the Macho-Nice Split"
  ) +
  theme_minimal(base_size = 8) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```
