---
title: "Macho-Nice Guy Dichotomy"
author: "Gilad Sarusi"
format: 
  html:
    toc: true
    number-sections: true
    self-contained: true
execute:
  warning: false
  message: false
editor: visual
---

# HistWords repo

For our purposes, I use the same approach chosen by Tessa E. S. Charlesworth and colleagues—namely, the HistWords repository, which is available on [GitHub](https://github.com/williamleif/histwords?tab=readme-ov-file).

### What's going on?

HistWords is an open-source dataset and toolkit that provides word embeddings for different historical periods, trained on Google Books Ngram (and other corpora). Each decade (1800–1990) has its own 300-dimensional vector space representing the meaning of words in that era.

Think of it as a longitudinal semantic dataset—rather than tracking people over time, we track words across decades. Each vector encodes a word’s position in the shared “meaning space” of its time.

#### What can we do with it?

- **Inspect meaning:** find the nearest neighbors of a word in any decade.

- **Compare meanings over time:** compute cosine similarity for the same word across decades to track semantic drift.

- **Study cultural change:** analyze how a word’s associations (e.g., *macho*, *nice*) evolve by decade.

- **Compose phrases:** average vectors (e.g., *macho* + *guy*) to represent multi-word concepts.

- **Align embeddings:** apply sequential Procrustes alignment to make spaces from different decades comparable.

### What are we comparing?

In this project, I adopt an approach similar to Tessa et al. designed to avoid the **alignment problem** inherent in diachronic word embeddings—which makes direct comparison of raw cosine similarities between two embedding spaces (e.g., the 1910s and the 1920s) invalid.

Instead of comparing one decade directly with another, I compare words **within each decade**. In other words, for every decade I ask a simple question: is a given word closer to the *Macho* prototype or to the *Nice Guy* prototype?

To do this, the code builds a vector for each prototype in each decade and then gives every word a **single contrast score**. Positive scores mean the word is closer to the Macho side; negative scores mean it is closer to the Nice side. This lets me describe change across time without pretending that the raw geometry of one decade can be directly compared with the raw geometry of another.

More concretely, the current script does four things:

1.  it builds decade-specific *Macho* and *Nice Guy* vectors in two different ways: a **nudge** version and an **inject** version;
2.  it uses an anchored scoring method to estimate where each word falls on the Macho-versus-Nice contrast within that decade;
3.  it keeps only words that also appear in the Warriner ratings, so later comparisons can be tied to valence; and
4.  in the Attractiveness section, it compares two scoring workflows side by side: the original pairwise cosine approach and the newer anchored approach.

Nevertheless, following Tessa et al.’s guidance, applying **orthogonal Procrustes** rotation can still be useful as a robustness check or for future extensions involving cross-decade word-trajectory analyses.

::: callout-cautio
`The following code is written using the ds4psych: Data Science for Psychology: Natural Language workflow (https://ds4psych.com). Therefore, the toolset used here differs from Tessa’s.`{.smallcaps}
:::

```{r loadPackages, echo = FALSE, message=FALSE}
# ----------------------------------------------------
#Packages
library(tidyverse)
library(vroom)       # fast reading of large texts
library(stringr) # string manipulation
library(purrr) # functional programming
library(sweater) # for loading embeddings
library(quanteda.textmodels) # PLS
library(quanteda) # for text analysis
library(embedplyr) # for working with embeddings manipulation
library(knitr) # kable
library(pls) # PLS


```

```{r Datapaths, include=FALSE}

# --- 0) Paths  -------------------

decades <- seq(1800, 1990, by = 10)      
model_names <- setNames(paste0("eng.all_sgns.",decades), decades) # all English books (Google N-grams)


```

Notably, in the 2000s there was a change in N-gram sampling. Therefore, embeddings using the corpus from the 2000s onward are excluded.

```{r}
## Words norm - for valence computation: 


warriner_path <- "wordstim/allwrdnorms_warriner.csv"         
# columns: word, valence, arousal, dominance - valence is the average rating on a 9-point scale from 1 (very negative) to 9 (very positive), obtained from Warriner and colleagues (2013)

# Theoretically we need to cosider that the fact that this was a crowdslabeling procedure taken place in 2013, the chatechristics of each trait in term of valence, arousal and dominance are reflecting the meaning of the word in 2013, which might be different from the meaning of the word in 1800. However, since we are interested in the relative valence of traits (e.g., is "kind" more positive than "dominant"?), and since these relative valence ratings are likely to be more stable over time than absolute ratings, I will proceed with using Warriner's ratings as a proxy for trait valence across all decades. This is a common approach in historical semantics research, although it is important to acknowledge this limitation when interpreting results.

```

As an alternative, we might consider NRC VAD (S. M. Mohammad, 2018), which contains 20,007 words with ratings between 0 and 1 for valence, arousal, and dominance.

```{r}

## List of traits:

traits_path <- "wordstim/traitlist.txt"

# We would latter on create a file storing those words for the code to be more elegant :

nice_guy_words <- c(
  "kind", "gentle", "caring", "sensitive", "considerate", "respectful",
  "supportive", "understanding", "warm", "affectionate",
  "polite", "reliable", "faithful", "honest", "thoughtful",
  "attentive", "humble", "nice","whipped", "romantic"
)
macho_guy_words <- c(
   "dominant", "assertive", "confident", "tough", "strong", "decisive",
  "bold", "powerful", "competitive", "aggressive", "independent", "unemotional",
  "fearless","alpha", "commanding", "ambitious", "macho"
)

```

### Two ways of defining the constructs

To make sure these trait lists are interpreted in a masculine context, I use two slightly different strategies.

The first is to **inject male anchors**. This means I add words such as *man*, *men*, *male*, and *guy* to each list before averaging the embeddings. The goal is to pull the resulting vector toward a masculine semantic context rather than a generic trait context. This is implemented in the *Concepts* chunk.

- **Cons / watch-outs:**\
  *Anchor dominance risk*—male anchors can outweigh trait terms if the lists are small. Moreover, shared anchors in both groups reduce the distance between group centroids (smaller margins).\
  **Monitor:** anchor share in each centroid (≤30–40%), cosine margin between group means, and Top-N stability across decades (overlap %).

The second approach, implemented in the next chunk, is a **masculinity projection nudge**. Here I build a male-versus-female direction in the embedding space and then shift each construct slightly toward the male side before scoring it. This is a softer adjustment than injection: instead of adding anchor words into the list itself, it gently pushes the final vector toward a masculine interpretation.

- **Cons / watch-outs:**\
  *Axis quality*—if the gender sets are unbalanced or drift strongly by decade, the direction can be noisy.\
  *Leakage*—over-weighting (large *w*) can wash out original semantics.\
  **Monitor:** axis stability across decades (cosine between decade-specific *d*<sub>male</sub>), use small *w* (e.g., 0.1–0.3) and run sensitivity analyses; compare results with and without the tweak.

Using both approaches gives me a built-in robustness check. If the results tell the same story under both methods, that makes the interpretation more convincing.

### Masculinity projection nudge {#musculinity-nudge}

To interpret these trait bundles as referring to men rather than to personality in general, I create a female-to-male direction in the semantic space and add part of that direction to the construct vector. In plain language, this step nudges the vector so that later comparisons are more likely to reflect the masculine reading of the traits.

```{r}



# E: numeric matrix (rows = tokens, cols = dims), 

# male_tokens / female_tokens: anchors for the axis (can leave defaults)

 male_anchors <-  c("man","men","male","guy")  #Question: should we add 'boy'? 
 female_anchors <-  c("woman", "women", "female","girl")

 
 build_male_female_axis <- function(embeddings) {
  male_ddr <- predict(embeddings, male_anchors) |> average_embedding()
  female_ddr <- predict(embeddings, female_anchors) |> average_embedding()
  male_ddr - female_ddr
}

 

# DDR-based masculinity function:

# w = nudge strength (0 = no nudge, >0 = push toward masculine direction)
 
# default is set to 1 to inteperate results easily and to make sure that the wors be are assigning is regards to word_GUY not that word in general
 
 
masculinity_nudge <- function(ddr_vec, # Or any dim1:dim300 embedding object
                              E, # The embedding space 
                              w = 1) {
 axis <- build_male_female_axis(E)
  axis <- as.numeric(axis)
  
  if (is.matrix(ddr_vec)) {
    
    ddr_vec_nudged <- sweep(
      ddr_vec,
      2,
      w * axis,
      FUN = "+"
    )
    
  } else {
    
    ddr_vec <- as.numeric(ddr_vec)
    ddr_vec_nudged <- ddr_vec + w * axis
    
  }
  
  ddr_vec_nudged

}



```

The second approach, the **Inject approach**, does the same job more directly: it adds male anchor words to each word list before averaging, so the resulting construct vector is explicitly tied to a masculine context.

```{r Concepts, include=FALSE}
# --- 2) Defining concepts (labels) -----------------------------


# Inject them inside the Macho/Nice Construct 
nice_guy_words_injected <- nice_guy_words |>
  c(male_anchors)

macho_guy_words_injected <- macho_guy_words |>
    c(male_anchors)

```

```{r loadData1, include=FALSE}
# --- 3) Loading Data  ----------------


## -------------  Warriner's valence-----------

#(I'm keeping Arousal and Dominance since it might be useful later)

warriner <- read_csv(warriner_path, show_col_types = FALSE) |>
  transmute(word=tolower(Word),
         #Valence
         valence=V.Mean.Sum,
         # Arousal
         arousal=A.Mean.Sum,
         # Dominance
         dominance = D.Mean.Sum,
         id = as.character(row_number())
         ) |>
  
  # dealing with duplicate: averageing across duplicate
  group_by(word) |>
  summarise(
    valence = mean(valence, na.rm = TRUE),
    arousal = mean(arousal, na.rm = TRUE),
    dominance = mean(dominance, na.rm = TRUE),
    id = first(id)
  ) 


## ------------ Lists of traits --------------


trait_list<- read.delim(traits_path, header = FALSE) |>
  # the traits are stored in a column named V1 and I want to use it as a vector 
  pull(V1) 


  # I would fetch only the traits that got crowdslebelling in warriner: 
traits_warriner <- warriner |>
  filter(word %in% trait_list) 
  
```

### ENGall Model

Properties of ENGall_model:

- n of words: 100,000 (same word list across all decades). the authors only computed embeddings for words that were among the top 100,000 most frequent words over all time (for EngAll)
  - n of dimensions: 300.
  - Type of Embedding : SGNS.
  - Context Window : symmetric context window of 4 words.

```{r}
## ----------- ENGall model -----------

if (file.exists("engall/ENGall_model.RDS")) {
  ENGall_model<-readRDS("engall/ENGall_model.RDS")
}else {
  ENGall_model <- lapply(model_names, load_embeddings, save = FALSE) # list of embedding models
  saveRDS(ENGall_model, 
          file="engall/ENGall_model.RDS")
}
length(ENGall_model)
```

The ENGall model provides embeddings for the 100,000 most frequent words **across** all time. This means words coined later in history will be missing in earlier slices, and some words will be absent in particular decade models. The model is stored as a list of 20 elements, one per decade. In some decade data frames, certain words have no embedding.

let's map those words.

```{r}

#### ---- Check unavailable words by decade ----
n_avwords <- vector()
n_unav<- vector()
prop_avail <- vector()
n_total <- vector()

# Creating a list of available word and unavailable words: 

for (i in seq_along(ENGall_model)) {
  # Count available words as those with V1 != 0
  n_avwords[i] <- sum(ENGall_model[[i]][, 1] != 0)
  n_total[i] <- ENGall_model[[i]]|>
      nrow()  
  n_unav[i]  <- n_total[i] - n_avwords[i]
  prop_avail[i] <- n_avwords[i] / n_total[i]
}


avwords_n_decade_df <- data.frame(
  decade      = decades,
  n_avwords   = n_avwords,
  n_unavwords = n_unav,
  prop_avail  = prop_avail
)

```

```{r availWordsTable}
kable(avwords_n_decade_df |>
        #adding comma
        mutate(across(where(is.numeric),scales::comma)),
      digits = 3,
      caption = "Proportion of Words with Embedding")

```

Out of 100,000 words, each decade has on average 36,347 available words.

Notably, the 1990s slice has \~71% of words available. This does **not** mean “29% are post-1990 words.” Rather, 29% of the top-100k (over 1800–1999) simply lack enough occurrences in the 1990s to train vectors—often because they are earlier-era words that faded out, or because they are rare or orthographically different in that decade.

Later, we will pull valence from Warriner’s list (\~14,000 words), which bounds our effective vocabulary to those items. Because not all Warriner words appear in ENGall, the intersection may be smaller.

Another concern is whether some of the words used to define our constructs are missing from the model.

Let's check!

```{r}

### ---- Check if macho/nice words by decade ------- 
# 1) Availability (token x decade)

avail <- map2_df(ENGall_model, decades, ~
  tibble(
    decade    = .y,
    token     = tolower(rownames(.x)),
    available = .x[, 1] != 0
  )
)

#  Define the sets you care about
sets <- list(
  macho = macho_guy_words,
  nice  = nice_guy_words,
  male = male_anchors
)

wanted <- enframe(sets, name = "set", value = "token") |>
  unnest(token) |>
  mutate(token = tolower(token))

# proportion per decade + list of missing words 
macho_nice_words_by_decade_df <- wanted |>
  left_join(avail, by = "token") |>
  mutate(available = tidyr::replace_na(available, FALSE)) |>
  group_by(set, decade) |>
  summarise(
    n_of_words_examined = n(),
    n_available = sum(available),
    prop_avail    = n_available / n_of_words_examined,
    missing     = list(token[!available]),
    .groups = "drop"
  ) |>
  arrange(set, decade)

```

```{r}

kable(macho_nice_words_by_decade_df,
      caption= "Construct's word presence across time")
```

It seems the words *cocky* and *macho* do not appear in ENGall at all. I therefore omit them so that the construct vectors are built only from words that actually exist in the model.

This leaves the Macho and Nice lists with slightly different lengths, which is worth keeping in mind when interpreting the results.

```{r}
length(macho_guy_words)
#[1] 20

length(macho_guy_words_injected)

#[1] 24

length(male_anchors)
# [1] 4

macho_guy_words <- setdiff(macho_guy_words,c("cocky", "macho"))
macho_guy_words_injected <- setdiff(macho_guy_words_injected,c("cocky", "macho"))

# Sanity check
length(macho_guy_words)
# [1] 18

length(macho_guy_words_injected)

# [1] 20

# OK I'm sane
    
    
```

# DDR


I next build the decade-specific semantic representations for *Macho* and *Nice Guy*. In this document I use the term **DDR** simply to mean the vector that summarizes one construct in one decade.

The code creates four such vectors for each decade:

- Macho, using the nudge approach
- Nice Guy, using the nudge approach
- Macho, using the inject approach
- Nice Guy, using the inject approach

## Pairwise

The first scoring method is the *Pairwise Cosine-Delta* method. For every word in the decade-specific embedding space, cosine similarity is computed separately to the Macho DDR and to the Nice DDR. The final score is then defined as:

cosine_delta = cosine(word, Macho DDR) - cosine(word, Nice DDR)

Thus, positive values indicate that a word is more similar to the *Macho than to the Nice*, whereas negative values indicate that a word is more similar to the *Nice than to the Macho*. This method treats the two constructs as two separate reference points and asks whether a word is relatively closer to one than the other- non Dichatomic.

## Anchored Method

The second scoring method is the anchored Macho–Nice axis. Here, the Macho DDR is passed as the positive anchor and the Nice DDR is passed as the negative anchor using

`get_sims(..., method = "anchored").` Rather than computing two independent similarities and subtracting them, the anchored method places each word on a bipolar semantic continuum defined by the Nice pole at one end and the Macho pole at the other. Conceptually, this method asks where each word falls on the semantic axis:

Nice ←———— word ————→ Macho

This scale is ranging form 0 (nice) to 1 (macho). Higher anchored scores indicate a position closer to the Macho pole, whereas lower scores indicate a position closer to the Nice pole. This method therefore treats Macho and Nice as two ends of a single semantic dimension and therefore dichatomic.

## PLS
The third scoring method is the PLS-based Macho–Nice axis. In this approach, a supervised Partial Least Squares model is trained to distinguish Macho seed words from Nice seed words. Macho words are coded as +1, and Nice words are coded as -1. The resulting one-component PLS projection is used as a supervised semantic axis that maximally separates the two word groups. Each word in the decade-specific embedding space is then scored by its cosine similarity to this PLS-derived axis. Positive PLS scores indicate alignment with the Macho-oriented side of the supervised axis, whereas negative scores indicate alignment with the Nice-oriented side.

The three methods therefore answer related but non-identical questions. The pairwise cosine-delta method asks whether a word is more similar to Macho than to Nice. The anchored method asks where a word lies on the bipolar Nice–Macho continuum. The PLS method asks how strongly a word aligns with a supervised axis learned from the Macho versus Nice seed-word distinction. Because these methods produce scores using different geometric procedures, their raw magnitudes should not be interpreted as directly equivalent.

Instead, they should be compared in terms of directional agreement, rank-order similarity, overlap in high-scoring and low-scoring words, and associations with external lexical variables.
```{r pls-helper}
make_pls_macho_nice_axis <- function(embeddings,
                                     macho_words,
                                     nice_words,
                                     masculine_nudge = FALSE) {
  
  train_words <- tibble(
    word = c(macho_words, nice_words),
    y = c(
      rep(1, length(macho_words)),
      rep(-1, length(nice_words))
    )
  ) |>
    distinct(word, .keep_all = TRUE) |>
    filter(word %in% rownames(embeddings))

  X <- predict(embeddings, train_words$word)

  if (masculine_nudge) {
    X <- masculinity_nudge(X, E = embeddings)
  }

  X <- as.matrix(X)

  keep <- complete.cases(X) & rowSums(X^2, na.rm = TRUE) > 0

  X <- X[keep, , drop = FALSE]
  y <- train_words$y[keep]

  if (length(y) != nrow(X)) {
    stop("PLS training mismatch: X rows and y labels do not match.")
  }

  if (length(unique(y)) < 2) {
    stop("Only one class remains after removing missing embeddings.")
  }

  pls_data <- as.data.frame(X)
  pls_data$y <- y

  pls_fit <- plsr(
    y ~ .,
    data = pls_data,
    ncomp = 1,
    scale = FALSE,
    validation = "none"
  )

  axis <- as.numeric(pls_fit$projection[, 1])
  scores <- as.numeric(X %*% axis)

  if (mean(scores[y == 1]) < mean(scores[y == -1])) {
    axis <- -axis
  }

  axis / sqrt(sum(axis^2))
}
```

```{r DDRs,warning=FALSE}

# PlS
pls_nudge_axis_by_decade <- vector("list", length(ENGall_model))
pls_inject_axis_by_decade <- vector("list", length(ENGall_model))

Macho_DDR_nudge_by_decade <- vector("list", length(ENGall_model))
Nice_DDR_nudge_by_decade <- vector("list", length(ENGall_model))

Macho_DDR_injected_by_decade <- vector("list", length(ENGall_model))
Nice_DDR_injected_by_decade <- vector("list", length(ENGall_model))

decade_scores <- vector("list", length(ENGall_model))

for (i in seq_along(ENGall_model)) {
  embeddings <- ENGall_model[[i]]
  embeddings <- embeddings[magnitude(embeddings) != 0, ]
  
  # Nudged approach
  Nice_DDR_nudge_by_decade[[i]] <- masculinity_nudge(
    ddr_vec = average_embedding(predict(embeddings, nice_guy_words)),
    E = embeddings
  ) 
  
  Macho_DDR_nudge_by_decade[[i]] <- masculinity_nudge(
    ddr_vec = average_embedding(predict(embeddings, macho_guy_words)),
    E = embeddings
  ) 
  
  # Inject approach
  Nice_DDR_injected_by_decade[[i]] <- average_embedding(
    predict(embeddings, nice_guy_words_injected)
  )
  Macho_DDR_injected_by_decade[[i]] <- average_embedding(
    predict(embeddings, macho_guy_words_injected) 
  )
  
  # Method 1: Pairwise Deltas 
  cosine_scores_word_level <- embeddings |>
    get_sims(
      list(
        "macho_cos_nudge" = Macho_DDR_nudge_by_decade[[i]],
        "nice_cos_nudge" = Nice_DDR_nudge_by_decade[[i]],
        "macho_cos_inject" = Macho_DDR_injected_by_decade[[i]],
        "nice_cos_inject" = Nice_DDR_injected_by_decade[[i]]
      ), 
      method = "cosine"
    ) |> 
    mutate(
      cosine_delta_nudge = macho_cos_nudge - nice_cos_nudge,
      cosine_delta_inject = macho_cos_inject - nice_cos_inject
    ) |> 
    select(
      doc_id,
      cosine_delta_nudge,
      cosine_delta_inject
    ) 
  
  # Method 2: Anchored Macho-Nice Axis
  anchored_scores_word_level <- embeddings |> 
    get_sims(
      list(
        "macho_vs_nice_nudge" = list(
          pos = Macho_DDR_nudge_by_decade[[i]],
          neg = Nice_DDR_nudge_by_decade[[i]]
        ), 
        "macho_vs_nice_inject" = list(
          pos = Macho_DDR_injected_by_decade[[i]],
          neg = Nice_DDR_injected_by_decade[[i]]
        ) 
      ), 
      method = "anchored"
    )
  # Method 3: PLS
    pls_nudge_axis_by_decade[[i]] <- make_pls_macho_nice_axis(
    embeddings = embeddings,
    macho_words = macho_guy_words,
    nice_words = nice_guy_words,
    masculine_nudge = TRUE
  )
    
    pls_inject_axis_by_decade[[i]] <- make_pls_macho_nice_axis(
      embeddings = embeddings,
      macho_words = macho_guy_words_injected,
      nice_words = nice_guy_words_injected,
      masculine_nudge = FALSE
    )
    
    pls_scores_word_level <- embeddings |>
      get_sims(
        list(
          "pls_nudge"  = pls_nudge_axis_by_decade[[i]],
          "pls_inject" = pls_inject_axis_by_decade[[i]]
        ),
        method = "cosine"
      ) 
    
  decade_scores[[i]] <- anchored_scores_word_level |>
    left_join(cosine_scores_word_level,
              by = "doc_id")|>
    left_join(
      pls_scores_word_level |>
        select(
          doc_id,
          pls_nudge,
          pls_inject
        ),
      by = "doc_id"
    )
}

# Make it readable
names(Macho_DDR_nudge_by_decade) <- decades
names(Nice_DDR_nudge_by_decade) <- decades
names(Macho_DDR_injected_by_decade) <- decades
names(Nice_DDR_injected_by_decade) <- decades
names(decade_scores) <- decades


```

## Warriner List {#warriner}

Next, I keep **only** the words that also appear in the Warriner norms, because those are the words for which I have valence ratings.

In practice, this means I take the full set of anchored scores from the embedding model and trim it down to the subset of words that can later be interpreted in emotional terms.

```{r WarrinerOnly)}
# Starting with empty 20 long list 

warriner_with_scores <- vector("list", length(decade_scores))

for (i in seq_along(decade_scores)) {

  ds <- decade_scores[[i]]

  # --- standardize the token column name to "word" ---
  
  if ("doc_id" %in% names(ds)) {
    ds2 <- ds |> rename(word= doc_id)
  } else {
    stop(
      "Could not find a word column in decade_scores[[i]]. Check names(decade_scores[[i]])."
      )
  }

  # --- join: keep only Warriner words that exist in this decade vocab ---
  warriner_with_scores[[i]] <-
    warriner |>
    dplyr::inner_join(
      ds2,
      by = "word"
    )
}
names(warriner_with_scores) <- decades

warriner_with_scores


#-------- NOTICE! ------

# Before the comparisons I'll omit the words that were used to create the DDR, namely, macho_guy_word and nice_guy_words, to reduce noise .

omit_words <- unique(c(
  macho_guy_words, nice_guy_words,
  macho_guy_words_injected, nice_guy_words_injected
))

warriner_scores_omit <- lapply(warriner_with_scores, function(df) {
  df |> filter(!word %in% omit_words)
})


```



# RQs
## Q1

**Is there, overall, a difference in association across 200 years between Macho and Nice Guy?**

Here, positive values mean relatively stronger Macho association and negative values mean relatively stronger Nice association. So here we'll see whether the delta, anchored pole loading and pls changes, first overall and then by decade.


```{r compare-methods-prepare-long}
method_scores_long <- bind_rows(warriner_scores_omit, .id = "decade") |>
  mutate(
    decade = factor(decade, levels = as.character(decades)),
    decade_num = as.numeric(decade),

    # Anchored scores from get_sims(method = "anchored") are 0-1:
    # 0 = Nice pole, .5 = midpoint, 1 = Macho pole.
    # Center them so 0 = midpoint, positive = Macho, negative = Nice.
    anchored_nudge  = 2 * macho_vs_nice_nudge - 1,
    anchored_inject = 2 * macho_vs_nice_inject - 1,

    pairwise_nudge  = cosine_delta_nudge,
    pairwise_inject = cosine_delta_inject,

    pls_nudge  = pls_nudge,
    pls_inject = pls_inject
  ) |>
  select(
    word, id, valence, arousal, dominance,
    decade, decade_num,
    anchored_nudge, anchored_inject,
    pairwise_nudge, pairwise_inject,
    pls_nudge, pls_inject
  ) |>
  pivot_longer(
    cols = c(
      anchored_nudge, anchored_inject,
      pairwise_nudge, pairwise_inject,
      pls_nudge, pls_inject
    ),
    names_to = "score_type",
    values_to = "score"
  ) |>
  mutate(
    method = case_when(
      str_detect(score_type, "^anchored") ~ "Anchored",
      str_detect(score_type, "^pairwise") ~ "Pairwise delta",
      str_detect(score_type, "^pls") ~ "PLS"
    ),
    approach = case_when(
      str_detect(score_type, "nudge$") ~ "Nudge",
      str_detect(score_type, "inject$") ~ "Inject"
    ),
    direction = case_when(
      score > 0 ~ "Macho",
      score < 0 ~ "Nice",
      TRUE ~ "Neutral"
    ),
    valence_centered = valence - 5
  ) |>
  filter(is.finite(score))

```


```{r compare-methods-helper}
one_sample_score_summary <- function(x) {
  x <- x[is.finite(x)]
  
  n <- length(x)
  mean_score <- mean(x)
  sd_score <- if (n > 1) sd(x) else NA_real_
  se <- if (n > 1) sd_score / sqrt(n) else NA_real_
   
  if (n > 1 && is.finite(sd_score) && sd_score > 0) {
    test <- t.test(x, mu = 0)
     
    tibble(
      n = n,
      mean_score = mean_score,
      sd_score = sd_score,
      se = se,
      t_value = unname(test$statistic),
      df = unname(test$parameter),
      p_value = test$p.value,
      d_z = mean_score / sd_score,
      ci_low = unname(test$conf.int[1]),
      ci_high = unname(test$conf.int[2])
    )
  } else { 
    tibble(
      n = n,
      mean_score = mean_score,
      sd_score = sd_score,
      se = se,
      t_value = NA_real_,
      df = NA_real_,
      p_value = NA_real_,
      d_z = NA_real_,
      ci_low = NA_real_,
      ci_high = NA_real_
    )
  }
} 
```
### Word-level analysis

This asks: for each word, is its average score across decades reliably above or below zero?

```{r compare-methods-word-level}
compare_nice_macho_words <- function(data, only_traits = FALSE) {

  if (only_traits) {
    data <- data |>
      filter(word %in% traits_warriner$word)
  }

  data |>
    group_by(method, approach, word) |>
    group_modify(~ one_sample_score_summary(.x$score)) |>
    ungroup() |>
    left_join(
      data |>
        select(word, valence, arousal, dominance) |>
        distinct(),
      by = "word"
    ) |>
    mutate(
      direction = case_when(
        mean_score > 0 ~ "Macho",
        mean_score < 0 ~ "Nice",
        TRUE ~ "Neutral"
      ),
      p_fdr = p.adjust(p_value, method = "fdr"),
      sig = case_when(
        is.na(p_fdr) ~ "No",
        p_fdr < 0.001 ~ "***",
        p_fdr < 0.01 ~ "**",
        p_fdr < 0.05 ~ "*",
        TRUE ~ "No"
      )
    ) |>
    arrange(method, approach, p_fdr)
}

word_summary_all <- compare_nice_macho_words(
  method_scores_long,
  only_traits = FALSE
)

word_summary_traits <- compare_nice_macho_words(
  method_scores_long,
  only_traits = TRUE
)

word_summary_all |>
  filter(sig != "No") |>
  group_by(method, approach, direction) |>
  arrange(p_fdr, .by_group = TRUE) |>
  slice_head(n = 10) |>
  ungroup() |>
  select(method, approach, direction, word, n, mean_score, d_z, p_value, p_fdr, sig) |>
  kable(
    caption = "Top word-level Macho/Nice scores across decades by method and approach",
    digits = 3
  )

```
### Decade-level analysis

This asks: within each decade, are Warriner words on average shifted toward Macho or Nice?

```{r compare-methods-decade-level}
compare_overall_by_decade <- function(data, only_traits = FALSE) {

  if (only_traits) {
    data <- data |>
      filter(word %in% traits_warriner$word)
  }

  data |>
    group_by(method, approach, decade) |>
    group_modify(~ one_sample_score_summary(.x$score)) |>
    ungroup() |>
    mutate(
      direction = case_when(
        mean_score > 0 ~ "Macho",
        mean_score < 0 ~ "Nice",
        TRUE ~ "Neutral"
      ),
      sig = case_when(
        is.na(p_value) ~ "No",
        p_value < 0.001 ~ "***",
        p_value < 0.01 ~ "**",
        p_value < 0.05 ~ "*",
        TRUE ~ "No"
      )
    ) |>
    arrange(method, approach, decade)
}

decade_summary_all <- compare_overall_by_decade(
  method_scores_long,
  only_traits = FALSE
)

decade_summary_traits <- compare_overall_by_decade(
  method_scores_long,
  only_traits = TRUE
)

decade_summary_all |>
  select(
    method, approach, decade, n,
    mean_score, d_z, ci_low, ci_high,
    t_value, df, p_value, direction, sig
  ) |>
  kable(
    caption = "Decade-level Macho/Nice shift across Warriner words by method and approach",
    digits = 3
  )


```


### Overall method-level summary

This gives one result per `method × approach` across all word-decade observations.

```{r compare-methods-overall}
overall_summary_all <- method_scores_long |>
  group_by(method, approach) |>
  group_modify(~ one_sample_score_summary(.x$score)) |>
  ungroup() |>
  mutate(
    direction = case_when(
      mean_score > 0 ~ "Macho",
      mean_score < 0 ~ "Nice",
      TRUE ~ "Neutral"
    ),
    sig = case_when(
      is.na(p_value) ~ "No",
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      TRUE ~ "No"
    )
  )

overall_summary_all |>
  kable(
    caption = "Overall Macho/Nice shift across all Warriner word-decade observations",
    digits = 3
  )
```
## Q2

**Which traits are most consistently pulled toward Macho and Nice Guy across decades?**

To keep this interpretable, I focus here on the rated trait words rather than the full vocabulary. The goal is simple: average each trait's score across decades, keep the significant trait-level results, and then inspect the strongest positive effects (Macho) and strongest negative effects (Nice).

```{r compare-methods-top-traits}
apa_num <- function(x, digits = 2) {
  ifelse(
    is.na(x),
    NA_character_,
    sprintf(paste0("%.", digits, "f"), x)
  )
}

apa_p <- function(p) {
  case_when(
    is.na(p) ~ NA_character_,
    p < 0.001 ~ "< .001",
    TRUE ~ sub("^0\\.", ".", sprintf("%.3f", round(p, 3)))
  )
}

apa_ci <- function(low, high, digits = 2) {
  ifelse(
    is.na(low) | is.na(high),
    NA_character_,
    paste0(
      "[",
      sprintf(paste0("%.", digits, "f"), low),
      ", ",
      sprintf(paste0("%.", digits, "f"), high),
      "]"
    )
  )
}

top_signed_traits <- function(data, top_n = 10) {
  data |>
    filter(sig != "No") |>
    mutate(
      sort_score = if_else(direction == "Macho", d_z, -d_z)
    ) |>
    group_by(method, approach, direction) |>
    slice_max(order_by = sort_score, n = top_n, with_ties = FALSE) |>
    arrange(method, approach, direction, desc(sort_score), .by_group = TRUE) |>
    mutate(rank = row_number()) |>
    ungroup()
}

top_trait_summary <- top_signed_traits(word_summary_traits, top_n = 10)

prepare_trait_table <- function(data, focal_direction) {
  data |>
    filter(direction == focal_direction) |>
    transmute(
      Method = method,
      Approach = approach,
      Rank = rank,
      Trait = word,
      M = apa_num(mean_score, 2),
      `95% CI` = apa_ci(ci_low, ci_high, 2),
      d_z = apa_num(d_z, 2),
      `p (FDR)` = apa_p(p_fdr)
    )
}
```

Before the plot, I keep just the top five traits per method, approach, and direction. If the same trait appears in several method panels, I show it only once within the Macho or Nice row and leave the point visible only in the columns where that trait belongs to the top five. I still use a forest-style plot so the mean trait score and its 95% confidence interval are both visible.

```{r compare-methods-top-traits-plot}
top_trait_plot_data <- top_signed_traits(word_summary_traits, top_n = 5)

trait_row_order <- top_trait_plot_data |>
  group_by(direction, word) |>
  summarise(
    n_panels = n(),
    mean_priority = mean(sort_score, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(direction, desc(n_panels), desc(mean_priority), word) |>
  mutate(direction_word = paste(direction, word, sep = "___"))

top_trait_plot_data <- top_trait_plot_data |>
  left_join(
    trait_row_order |>
      select(direction, word, direction_word),
    by = c("direction", "word")
  ) |>
  mutate(
    direction_word = factor(
      direction_word,
      levels = rev(trait_row_order$direction_word)
    )
  )

ggplot(
  top_trait_plot_data,
  aes(x = mean_score, y = direction_word, color = direction)
) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_segment(
    aes(
      x = ci_low,
      xend = ci_high,
      y = direction_word,
      yend = direction_word
    ),
    linewidth = 0.8,
    show.legend = FALSE
  ) +
  geom_point(size = 2.4, show.legend = FALSE) +
  facet_grid(direction ~ method + approach, scales = "free_y", space = "free_y") +
  scale_y_discrete(labels = function(x) sub("^[^_]+___", "", x)) +
  scale_color_manual(values = c(Macho = "#B35C1E", Nice = "#1C7C7D")) +
  labs(
    title = "Top traits most consistently associated with each construct",
    x = "Mean score across decades with 95% CI",
    y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text = element_text(face = "bold"),
    panel.grid.major.y = element_blank()
  )
```

The tables below give the same result in a more readable format. I split them into Macho and Nice so each side of the contrast can be scanned on its own.

```{r compare-methods-top-traits-tables}
prepare_trait_table(top_trait_summary, "Macho") |>
  kable(
    caption = "Top Macho-leaning traits across decades",
    align = c("l", "l", "r", "l", "r", "l", "r", "l")
  )

prepare_trait_table(top_trait_summary, "Nice") |>
  kable(
    caption = "Top Nice-leaning traits across decades",
    align = c("l", "l", "r", "l", "r", "l", "r", "l")
  )
```

## Q3

**Over time, does the valence of the traits most associated with each construct change?**

Here I stay with traits only and work directly from the long data. Within each decade, I take the ten most Macho-leaning traits and the ten most Nice-leaning traits, compute their mean centered valence (`valence - 5`), and then track that mean over time. Zero is neutral, positive values are more positive, and negative values are more negative.

```{r compare-methods-top-trait-valence}
select_top_traits_by_decade <- function(data, top_n = 10) {
  data |>
    filter(
      word %in% traits_warriner$word,
      direction != "Neutral"
    ) |>
    mutate(
      rank_score = if_else(direction == "Macho", score, -score)
    ) |>
    group_by(method, approach, decade, decade_num, direction) |>
    slice_max(order_by = rank_score, n = top_n, with_ties = FALSE) |>
    ungroup()
}

top_trait_decade_scores <- select_top_traits_by_decade(
  method_scores_long,
  top_n = 10
)

top_trait_valence_by_decade <- top_trait_decade_scores |>
  group_by(method, approach, decade, decade_num, direction) |>
  summarise(
    n_traits = n(),
    mean_valence_centered = mean(valence_centered, na.rm = TRUE),
    sd_valence_centered = sd(valence_centered, na.rm = TRUE),
    se_valence = if_else(
      n_traits > 1,
      sd_valence_centered / sqrt(n_traits),
      NA_real_
    ),
    t_crit = if_else(
      n_traits > 1,
      qt(0.975, df = n_traits - 1),
      NA_real_
    ),
    ci_low = mean_valence_centered - (t_crit * se_valence),
    ci_high = mean_valence_centered + (t_crit * se_valence),
    mean_score = mean(score, na.rm = TRUE),
    .groups = "drop"
  )
```

The plot shows the decade-by-decade valence trajectory for those top traits under each method and approach. The shaded ribbons are 95% confidence intervals around the decade-specific mean valence.

```{r compare-methods-top-trait-valence-plot}
ggplot(
  top_trait_valence_by_decade,
  aes(
    x = decade_num,
    y = mean_valence_centered,
    color = direction,
    fill = direction,
    group = direction
  )
) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_ribbon(
    aes(ymin = ci_low, ymax = ci_high),
    alpha = 0.15,
    color = NA,
    show.legend = FALSE
  ) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.6) +
  facet_grid(approach ~ method) +
  scale_color_manual(values = c(Macho = "#B35C1E", Nice = "#1C7C7D")) +
  scale_fill_manual(values = c(Macho = "#B35C1E", Nice = "#1C7C7D")) +
  labs(
    title = "Mean valence of the top 10 traits within each decade",
    x = "Decade",
    y = "Mean centered valence",
    color = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "top"
  )
```

To summarize the time pattern more compactly, I fit a simple linear trend within each method, approach, and direction. The slope tells us whether the top traits become more positive or more negative across time.

```{r compare-methods-top-trait-valence-table}
summarize_valence_trend <- function(data) {
  model <- lm(mean_valence_centered ~ decade_num, data = data)
  model_summary <- summary(model)
  model_ci <- confint(model)["decade_num", ]
  start_row <- data |> arrange(decade_num) |> slice_head(n = 1)
  end_row <- data |> arrange(decade_num) |> slice_tail(n = 1)
  
  tibble(
    start_decade = start_row$decade,
    start_mean = start_row$mean_valence_centered,
    start_ci_low = start_row$ci_low,
    start_ci_high = start_row$ci_high,
    end_decade = end_row$decade,
    end_mean = end_row$mean_valence_centered,
    end_ci_low = end_row$ci_low,
    end_ci_high = end_row$ci_high,
    change_across_period = end_mean - start_mean,
    slope_per_decade = unname(coef(model)[["decade_num"]]),
    slope_ci_low = unname(model_ci[1]),
    slope_ci_high = unname(model_ci[2]),
    p_value = model_summary$coefficients["decade_num", "Pr(>|t|)"]
  )
}

top_trait_valence_trends <- top_trait_valence_by_decade |>
  group_by(method, approach, direction) |>
  group_modify(~ summarize_valence_trend(.x)) |>
  ungroup()

top_trait_valence_trends |>
  transmute(
    Method = method,
    Approach = approach,
    Direction = direction,
    Start = paste0(
      start_decade,
      ": ",
      apa_num(start_mean, 2),
      " ",
      apa_ci(start_ci_low, start_ci_high, 2)
    ),
    End = paste0(
      end_decade,
      ": ",
      apa_num(end_mean, 2),
      " ",
      apa_ci(end_ci_low, end_ci_high, 2)
    ),
    `Change in M` = apa_num(change_across_period, 2),
    `Slope b` = apa_num(slope_per_decade, 3),
    `95% CI for b` = apa_ci(slope_ci_low, slope_ci_high, 3),
    p = apa_p(p_value)
  ) |>
  kable(
    caption = "Linear change in the valence of the top 10 traits within each decade",
    align = c("l", "l", "l", "l", "l", "r", "r", "l", "l")
  )
```
## Attractiveness

This section intentionally keeps **both** Attractiveness workflows in the same document.

The older workflow, which came from `main`, asks two separate questions for each decade: "How close is Attractiveness to Macho?" and "How close is Attractiveness to Nice?" The newer workflow, from `anlysis-after-opt`, asks a single contrast question: "Is Attractiveness closer to Macho or to Nice?"

In principle, both approaches should tell a very similar story. Because they currently do not give identical results, I present them side by side and then compare them directly at the end.

### Word list

We begin by specifying a small list of adjectives that directly express attractiveness, such as *attractive*, *beautiful*, *handsome*, and *sexy*. As above, two versions are created:

- Nudge version: uses only the attractiveness descriptors.

- Inject version: includes the attractiveness descriptors plus the male anchor words, ensuring a stronger masculine reference frame in the embedding space.

This parallels the earlier Macho/Nice pipeline for consistency.

```{r attractiveness-words}
attractiveness_words <- c(
  "attractive",
  "beautiful",
  "handsome",
  "gorgeous",
  "pretty",
  "sexy",
  "cute",
  "alluring",
  "charming",
  "appealing"
)

attractiveness_words_injected <- attractiveness_words |>
  c(male_anchors)
```

For each decade, I build an Attractiveness vector using the same two strategies used above for Macho and Nice:

**Inject condition**

1.  Retrieve the embeddings of all attractiveness words + male anchors.
2.  Compute their **average embedding**, yielding the *Attractiveness Injected DDR* for that decade.

**Nudge condition**

1.  Retrieve the embeddings of the attractiveness adjectives.
2.  Average them to obtain the raw *Attractiveness Nudge DDR*.
3.  Apply the **`masculinity_nudge()`** transformation to align the DDR with the nudge paradigm used for the Macho/Nice DDRs.

```{r attractiveness-ddrs}
Attractiveness_DDR_injected_by_decade <- vector("list", length(ENGall_model))
Attractiveness_DDR_nudge_by_decade <- vector("list", length(ENGall_model))

for (i in seq_along(ENGall_model)) {
  embeddings <- ENGall_model[[i]]
  embeddings <- embeddings[magnitude(embeddings) != 0, ]
  
  Attractiveness_DDR_injected_by_decade[[i]] <- predict(
    embeddings,
    attractiveness_words_injected
  ) |>
    average_embedding()
  
  Attractiveness_DDR_nudge_by_decade[[i]] <- predict(
    embeddings,
    attractiveness_words
  ) |>
    average_embedding() |>
    masculinity_nudge(E = embeddings)
}

names(Attractiveness_DDR_injected_by_decade) <- decades
names(Attractiveness_DDR_nudge_by_decade) <- decades
```


::: callout-note
## What now

Stuff to need to take into consideration and to thourghly continue the analysis

method: 1- Currently the masculinity nudge is set to w=1, meaning that the the nudge is very very storng toward the masculine side. try maybe something more sable. \* should inspect show what happening when varying w

2- the Attractiveness DDR also is constructed using the same "masculine context" embedding both inject and nudge method (which is also set to w=1!) I suspect that the results showed in @warriner that almost all the words loaded greatly more to the macho side of the axis (as shown in *macho_vs_nice* columns) is due to that the attractiveness is also pushed too strongly to the masculinity side.

- should inspect what happend when attractiveness not in the "masculine side" non-inject and non-nudged

future:

3 - Attractiveness as a sexual properties- changing the word bank 4- euclidean distance between the macho and nice guy construct as a support of the existance of a split. \* since only relative distance is possible for comparison across decades, a method is needed to examing the change (distance ) between the constructs. I'm thinking of taking the delta distance of every word from our macho/nice constructs, compute a delta and average across ALL the words and see if this average delta changes across time/

5- what other axaminations would indicate such "stereo-type- Split"?
:::
